Intergenerational transport of double-stranded RNA limits heritable epigenetic changes

Version 1

Abstract

Abstract

RNAs in circulation carry sequence-specific regulatory information between cells in plant, animal, and host-pathogen systems. Such RNA can cross generational boundaries, as evidenced by somatic double-stranded RNA (dsRNA) in the nematode C. elegans silencing genes of matching sequence in progeny. Here we dissect the intergenerational path taken by dsRNA from parental circulation and discover that cytosolic import through the dsRNA importer SID-1 in the parental germline and/or developing progeny varies with developmental time and dsRNA substrates. Loss of SID-1 enhances initiation of heritable RNA silencing within the germline and causes changes in the expression of the sid-1-dependent gene sdg-1 that last for more than 100 generations after restoration of SID-1. The SDG-1 protein is enriched in perinuclear Z granules required for heritable RNA silencing but is expressed from a retrotransposon targeted by such silencing. This auto-inhibitory loop reveals how retrotransposons could persist by hosting genes that regulate their own silencing.

Main text

RNAs released into circulation can act as intercellular messages that are used for gene regulation in distant cells. Examples include secretion of small RNAs within exosomes in response to pathogenic fungal infection in Arabidopsis1, virus-like proteins with their coding mRNAs in developing Drosophila2 and mice3, microRNAs from adipose tissue in mice4 and small RNAs from the epididymis in mice5,6,7,8. Such extracellular RNAs have also been detected in humans, but their roles in gene regulation remain unclear despite their use as diagnostic markers for diseases (reviewed in ref.9). Furthermore, the recent development of double-stranded RNA (dsRNA)-based drugs (reviewed in ref.10; refs.11,12) that can silence genes of matching sequence through RNA interference13 has heightened interest in understanding the import of dsRNA into cells. A conserved dsRNA-selective importer, SID-114,15,16, is required for the import of extracellular dsRNA into the cytosol of any cell in the nematode C. elegans. This entry into the cytosol is distinct from and can follow the uptake of dsRNA into cells, which can rely on other receptors (e.g., SID-2 for uptake from the intestinal lumen17,18). SID-1 has two homologs in mammals – SIDT1 and SIDT2. Although similar cytosolic entry of dsRNA through these mammalian homologs of SID-1 is supported by studies in mice reporting entry of viral dsRNA through SIDT219, enhanced dsRNA uptake when SIDT1 is overexpressed in vitro20, and uptake of ingested dsRNA into cells through SIDT121, alternative roles for SIDT1 and/or SIDT2 in the uptake of cholesterol have also been proposed22.

Secretion of dsRNA from C. elegans tissues expressing dsRNA from transgenes has been inferred based upon the SID-1-dependent silencing of matching genes in other tissues14,23. Secreted dsRNA from neurons can silence genes of matching sequence in most somatic cells24 and within the germline25. Extracellular dsRNA delivered into parental circulation by injection or ingestion also enters the germline and can cause silencing of matching genes in progeny13,26,27,28,29. In every case, the entry of dsRNA into the cytosol dictates when and where the processing of extracellular dsRNA can begin. Such intergenerational transport of RNA is an attractive mechanism for explaining endogenous, gene-specific effects in progeny that could occur in response to changes in somatic tissues of parents. However, which conditions induce transport of dsRNA into the germline, when during development this transport occurs, and what regulatory consequences ensue in progeny upon uptake of extracellular dsRNA from parents are all unknown. Despite this lack of knowledge, the analysis of transgenerational gene silencing triggered by dsRNA has revealed that a class of small RNAs called 22G RNA made using the mRNA targeted by dsRNA and bound by Argonaute proteins in the germline30,31 is necessary for observing silencing in every generation.

Timed dsRNA injection into animals or timed dsRNA ingestion by animals has been the only way to study mechanisms of dsRNA transport throughout C elegans development since no tools to induce dsRNA secretion from cells have been developed. Injection of dsRNA into adult sid-1(-) animals demonstrated that extracellular dsRNA can be directly transmitted to progeny without entry into the cytosol28,29. This intergenerational transmission of dsRNA in sid-1(-) adult animals requires the yolk receptor RME-232 and is independent of parental 22G RNA production because dsRNA cannot enter the cytosol in the parent. In further support of the initial intergenerational transport of extracellular dsRNA or dsRNA-derived signals not requiring 22G RNAs, gfp-dsRNA from parents that lack gfp sequences are transported to progeny (Fig. 1E in ref.28). Thus, the silencing signals transported from parent to progeny include the extracellular dsRNA and its derived silencing signals, independent of 22G RNAs. However, 22G RNAs are required for silencing and they may be used for intergenerational transport in subsequent generations.

Gene silencing by ingested dsRNA during larval development does not persist into adulthood.

Silencing in the germline was measured after exposure of gtbp-1::gfp animals to bacteria expressing dsRNA by imaging separate cohorts at indicated stages (a) or day 3 of adulthood (b). a and b, left, Schematics depicting stages and durations of exposure to dsRNA. a and b, right, GFP intensity (a.u.) in gtbp-1::gfp animals quantified in germ cells (larvae) or eggs in utero (adults) (a) or in day 3 adult (a3) animals (b) after exposure to control dsRNA (black) or gfp-dsRNA (red). The numbers of animals scored at each stage (n) are indicated. Asterisks indicate P < 0.05 with Bonferroni correction using Mann-Whitney U test for two-sided comparisons between animals exposed to control or gfp-dsRNA. Also see Extended Data Fig. 1.

All dsRNAs, regardless of length, have been assumed to be equivalent substrates for entry into the cytosol. This assumption is supported by the uptake of a variety of dsRNA substrates when C. elegans SID-1 is overexpressed in heterologous Drosophila S2 cells16. Yet, two key observations suggest that dsRNA can take multiple routes in vivo before SID-1-dependent entry into the cytosol. One, even in the presence of SID-1, dsRNA ingested during early adulthood requires RME-2 to cause silencing in progeny28. Two, 50-bp fluorescently-labeled dsRNA requires RME-2 for entry from parental circulation into oocytes28. Establishing the contexts for the use of different modes of dsRNA transport is crucial for understanding the processes regulated by endogenous extracellular dsRNA.

Here, we dissect the intergenerational transport of extracellular dsRNA and discover a role for this mechanism in modulating RNA regulation within the germline. Extracellular dsRNA is transported with developmental and substrate specificity from parental circulation to progeny, and its release from neurons can be enhanced using light-induced oxidative damage. Blocking dsRNA import into the cytosol of all cells revealed heritable changes in gene expression and led to the identification of sid-1-dependent gene-1 (sdg-1). The sdg-1 coding sequence is located within a retrotransposon that is targeted by RNA silencing in the germline. Yet, the SDG-1 protein colocalizes with regulators of RNA silencing in perinuclear granules within the germline and dynamically enters the nucleus in proximal oocytes and in cells of developing embryos. Measurements of sdg-1 expression using native mRNA, a translational reporter, or a transcriptional reporter reveal that expression is easily perturbed in different mutants that impact dsRNA-mediated gene regulation. Expression varies between the two gonad arms of wild-type animals, and different mutant isolates can show an increase or decrease in expression, indicative of a loss of buffered gene expression within the germline. However, consistent with an overall role for SDG-1 (and potentially other SDGs) in promoting RNA silencing, either loss of SID-1 or overexpression of SDG-1 enhances piRNA-mediated silencing within the germline initiated by mating. Therefore, we propose that the import of extracellular dsRNA into the germline tunes intracellular pathways that cause heritable RNA silencing.

Requirements for the entry of extracellular dsRNA into the germline change during development

A convenient method for the delivery of extracellular dsRNA into C. elegans at various times during larval development is the expression of dsRNA in the bacteria that the animals ingest as food26. To determine when ingested dsRNA can enter the germline and cause silencing, we exposed developing animals with a ubiquitously expressed protein (GTBP-1) tagged with GFP to bacteria that express gfp-dsRNA. Silencing was detectable within the germline from the second larval stage (L2) onwards (Fig. 1a and Extended Data Fig. 1a), but either exposure to ingested dsRNA beyond the fourth larval stage (L4) (Fig. 1b) or injection of dsRNA into the 1-day old adult germline (Extended Data Fig. 1b) was required to observe silencing in the germline of 3-day old adults. Combined with the need for exposure to dsRNA after the L4 stage28,29 for silencing in progeny, even for just 24 hours (Extended Data Fig. 1c), these observations suggest that heritable RNA silencing is not effectively initiated during early development of the germline despite detectable silencing within the germline. One possible explanation for this observation could be that both RNAs derived from the imported dsRNA and downstream silencing signals are continually diluted by the proliferation of germ cells. Heritable silencing by dsRNA ingested from the L4 stage to the first day of adulthood likely relies on entry of dsRNA into the proximal germline because silencing of a somatic gene in progeny after parental ingestion of dsRNA during this period required RME-2 (Extended Data Fig. 1c), which is enriched in the proximal germline (Extended Data Fig. 1d and ref.32), and some gtbp-1::gfp animals exposed to gfp-dsRNA up to the first day of adulthood showed more silencing in the proximal germline (Extended Data Fig. 1e).

Thus, these results reveal three periods of germline development that can be broadly distinguished based on the response to ingested dsRNA: (1) from the first larval to the fourth larval stage when exposure to dsRNA does not result in maximal silencing within the germline in adults (Fig. 1b); (2) from the fourth larval stage to early adulthood when entry of dsRNA primarily occurs in the proximal germline through RME-2 (Extended Data Fig. 1c,e); and (3) later adulthood when germline silencing by ingested dsRNA is maximal (Fig. 1b) and ingested dsRNA can effectively silence progeny independent of RME-2 (Extended Data Fig. 1c and ref.29). These differences in the entry of ingested dsRNA into cells and/or subsequent silencing could be driven by a variety of changes during development. These include changes in the uptake of dsRNA into the intestine, distribution of dsRNA to other tissues from the intestine, import of dsRNA into the germline, and availability of RNA silencing factors within the germline.

Oxidative damage in neurons expressing dsRNA enhances silencing in the germline by neuronal dsRNA

Another approach for delivering extracellular dsRNA into the germline that better mimics dsRNA transport between cells is the secretion of dsRNA from neurons25. However, the extent of such secretion throughout development is unpredictable. To modulate the secretion of dsRNA from somatic cells into parental circulation during development, we adapted an optogenetic approach for damaging somatic cells33. Specifically, we generated animals that express the mini singlet oxygen generator (miniSOG) protein in neurons and exposed them to blue light. While animals expressing miniSOG from a single-copy transgene did not show an appreciable defect when compared with wild-type animals, those expressing miniSOG from a multi-copy transgene were paralyzed (Extended Data Fig. 2a,b, top) and had visibly damaged neurons (Extended Data Fig. 2b, bottom). Using this system, we induced oxidative damage in the neurons of animals that expressed dsRNA under the control of a neuronal promoter and evaluated silencing of target genes with matching sequence expressed in other tissues (Fig. 2a). By exposing animals to blue light for 60 minutes at different times during development (Extended Data Fig. 2c), we observed SID-1-dependent enhancement in the silencing of the hypodermal gene bli-1 at the adult stage by neuronal bli-1-dsRNA, with maximal silencing when oxidative damage occurred during mid-to-late larval development (Extended Data Fig. 2d, light exposure every 6 hours from 42 to 66 hours post L4-stage of parent; Extended Data Fig. 2e, ∼2-fold increase from 14.9% to 29.1% in a background with enhanced RNA interference (eri-1(-)) and ∼6-fold increase from ∼1.6% to ∼9.8% in a wild-type background). A similar period of maximal SID-1-dependent enhancement of silencing was also observed when neurons expressing gfp-dsRNA were damaged and silencing of a two-gene operon that expresses two fluorescent proteins, mCherry::H2B and GFP::H2B, in the germline was measured (Fig. 2b-d, and Extended Data Fig. 2f – 48 to 60 hours post L4-stage of parent; sid-1(-) allele (jam80[nonsense]) is depicted in Extended Data Fig. 3). While silencing of gfp::h2b was observed throughout the germline, silencing of the other cistron mCherry::h2b was also observed sometimes, albeit restricted to specific regions of the germline. Silencing of mCherry::h2b was most frequent in the proximal germline and was not observed in any other region without silencing in the proximal germline (proximal germline – 57%, distal germline – 47%, sperm – 29%, Fig. 2d), likely due to reduction of mCherry::h2b::gfp::h2b pre-mRNA34 in those regions. Consistently, the silencing of both gfp::h2b and mCherry::h2b was eliminated in the absence of the nuclear Argonaute HRDE-1 (Extended Data Fig. 2g). The pattern of mCherry::h2b silencing is similar to the spatial pattern observed for the RME-2-dependent entry of dsRNA delivered into parental circulation28 and is consistent with the pattern of target mRNA degradation in the germline by extracellular dsRNA35. However, silencing of gfp::h2b in the germline by neuronal dsRNA did not show a detectable dependence on RME-2 (Extended Data Fig. 2h - difficulty in obtaining transgenic animals that also lack RME-2 resulted in a low sample size for this experiment).

Oxidative damage of neurons enhances gene silencing by neuronal dsRNA in the adult germline.

a, Schematic illustrating exposure to blue light of animals expressing a singlet oxygen generator (miniSOG) and gfp-dsRNA in neurons, and subsequent release of dsRNA. Such extracellular dsRNA is expected to enter the cytosol of the germline through the dsRNA importer SID-1 and silence gfp::h2b mRNA from a two-gene operon that expresses mCherry::h2b and gfp::h2b as part of a single pre-mRNA. b-d, Images of single gonad arms in adult animals with the two-gene operon (mex-5p::mCherry::h2b::gfp::h2b) showing fluorescence (black) of mCherry::H2B (magenta outline) or of GFP::H2B (green outline). Punctate autofluorescence from the intestine can also be seen. Numbers of animals assayed (n) and percentages of adult animals with the depicted expression patterns are indicated. Scale bars, 50 μm. b, mCherry::H2B fluorescence is seen throughout the germline (left) and GFP::H2B fluorescence is seen in the oocytes and in the distal gonad (right). c, GFP::H2B fluorescence in sid-1(+) and sid-1(-) animals expressing membrane-localized miniSOG (PH::miniSOG) and gfp-dsRNA driven by a neuronal promoter (rgef-1p) from a multi-copy transgene (Ex, jamEx214) without (left) or with (right) exposure to blue light at 48 hours post L4-stage of parent. d, mCherry::H2B fluorescence in sid-1(+) animals with the transgene Ex. Silencing of mCherry is enhanced in the distal gonad (third row) and sperm (fourth row) after exposing animals to blue light at 48 hours and 54 hours post L4-stage of parent. By region, silencing after exposure to light (right) in the proximal germline (57% = 10 + 18 + 29) > distal germline (47% = 18 + 29) > sperm (29%). Also see Extended Data Fig. 2 and Extended Data Fig. 3.

Thus, by modulating the secretion of dsRNA from somatic cells for the first time, we gained two insights into the intercellular transport of dsRNA: (1) oxidative damage of neurons during particular periods in development increases the amount of dsRNA and/or changes the kinds of dsRNA in circulation either because of specific enhancement of secretion or nonspecific spillage; and (2) there is a preference for silencing by neuronal dsRNA in the proximal germline. However, the temporal and/or spatial preferences observed for silencing by both neuronal and ingested dsRNA could be because of unknown characteristics of the exported neuronal dsRNA or ingested dsRNA expressed from bacteria (e.g., modifications, lengths, structures, etc.) that influence import or subsequent silencing. This hypothesis is supported by the different genetic requirements reported for silencing by neuronal gfp-dsRNA compared to other sources of gfp-dsRNA24. Alternatively, these preferences could reflect universal constraints for silencing using any extracellular dsRNA (e.g. expression patterns of factors that promote the import or processing of dsRNA).

Extracellular dsRNA in parental circulation can be transported through multiple routes to cause silencing in progeny

While the characteristics of extracellular dsRNA imported into the germline from ingested bacteria or from neurons are unknown, delivery of chemically defined dsRNA into the extracellular space in C. elegans can be accomplished using microinjection13,28. We examined differences, if any, in the entry of in vitro transcribed dsRNA from the extracellular body cavity into the germline during the L4 and adult stages as evidenced by silencing in the progeny of injected animals. Silencing was comparable regardless of whether wild-type or rme-2(-) parents were injected as L4-staged or adult animals (Fig. 3a and Extended Data Fig. 4a, left; also reported for adults in ref.29). However, some dependence on RME-2 for silencing in progeny was discernable when lower concentrations of dsRNA were used (Extended Data Fig. 4a, right). This result and previous results demonstrating proximal germline-specific silencing in contexts where silencing is less robust (Extended Data Fig. 1e and Fig. 2d) are consistent with RME-2-dependent uptake of dsRNA being a route for extracellular dsRNA to enter the germline. The difference in parental RME-2 requirement for silencing in progeny observed for dsRNA ingested (Extended Data Fig. 1c) or injected (Extended Data Fig. 4a) at the L4 stage could similarly reflect the accumulation of different amounts of dsRNA in parental circulation (e.g., more upon injection than upon ingestion), and/or different kinds of dsRNA (e.g., because of modifications in bacteria or upon transit through the intestine). However, these possibilities could not be easily distinguished because sensitive northern blotting36 revealed that both bacterial and in vitro transcribed dsRNA consist of a complex mix of dsRNAs (Extended Data Fig. 4b-d; consistent with refs.37,38), hereafter called mixed dsRNA.

Transport of dsRNA from parental circulation to progeny occurs through two routes with distinct substrate selectivity.

a, Hermaphrodite animals of indicated genotypes (in red) were injected in the body cavity with 50-bp unc-22-dsRNA synthesized with a 5’-OH (short dsRNA, left bars) or unc-22-dsRNA with a 5’ triphosphate transcribed from a ∼1.1 kb template (mixed dsRNA, right bars). Hermaphrodite self-progeny of injected animals were scored for unc-22 silencing (fr. Unc-22: strong, black; weak, grey). Numbers of injected parents and scored progeny (P0; F1 n) are indicated. Also see Extended Data Fig. 3 and Extended Data Fig. 4. b, Fluorescence images of progeny from animals with a gfp tag of the ubiquitously expressed gene gtbp-1 (gtbp-1::gfp) that were not injected (left), injected with 50-bp gfp-dsRNA (short dsRNA injection, middle), or injected with dsRNA transcribed from a ∼730-bp template (mixed dsRNA injection, right). Complete silencing is not observed in neurons or in the developing vulva; brackets indicate additional regions with dim GFP fluorescence. Numbers of animals assayed (n) and percentages of L4-staged animals with the depicted expression patterns are indicated. Scale bar, 100 μm. Also see Extended Data Fig. 5. c, Requirements for intergenerational transport of extracellular dsRNA. (top left) Differential Interference Contrast (DIC) and fluorescence images of a developing embryo from an animal injected in the body cavity with 50-bp dsRNA of the same sequence as in b and labeled at the 5’ end of the antisense strand with Atto-565. Accumulation within the intestinal lumen (arrowhead), number of embryos imaged (n), and percentage of embryos with depicted pattern of fluorescence are indicated. Scale bar, 20 μm. (top right and bottom) Hermaphrodite animals of the indicated genotypes were injected with short dsRNA (left bars) or mixed dsRNA (right bars) and self-progeny (top right) or cross progeny after mating with wild-type males (bottom) were analyzed as in a. Cases of no observable silencing are indicated with ‘0’. d, Schematic summarizing requirements for transport of dsRNA from parental circulation to developing progeny. See text for details. Asterisks in a and c indicate P < 0.05 with Bonferroni correction using χ2 test.

In contrast, when synthesized gfp-dsRNA of a defined length (50 bp) with a fluorescent label was injected into circulation in adult animals, no entry into the germline was observed in the absence of RME-228. We found that silencing of unc-22 in progeny by similarly synthesized but unlabeled 50-bp unc-22-dsRNA with a 5’ OH delivered into parental circulation also showed a strong requirement for RME-2, unlike mixed dsRNA (Fig. 3a). Further comparison between the two forms of dsRNA revealed that silencing in progeny by 50-bp dsRNA injected into parental circulation was detectably less efficient in somatic cells (Fig. 3b and Extended Data Fig. 5a,b, left), even when ∼14X more 50-bp dsRNA was delivered into parental circulation (Extended Data Fig. 5b, right), and was also less efficient in the germline (Fig. 3b and Extended Data Fig. 5a,c). Efficient silencing in response to added dsRNA requires nuclear Argonaute proteins: NRDE-3 in somatic cells34 and HRDE-1 in the germline30. Both 50-bp dsRNA and mixed dsRNA relied on the HRDE-1 for silencing within the germline (Extended Data Fig. 5a,c) and could silence independent of NRDE-3 in somatic cells (Extended Data Fig. 5a,c). Therefore, the observed difference in the extent of silencing cannot be attributed to differential engagement of these Argonautes, but rather could be the result of differences in the ability of each type of dsRNA to bind to upstream factors in the RNA interference pathway (e.g. RDE-4), differences in the stability of each type of dsRNA, and/or differences in the intergenerational transport of each type of dsRNA. In addition to the diversity of RNA lengths observed in mixed dsRNA, another known feature that could distinguish dsRNA transcribed in bacteria or in vitro from synthesized 50-bp dsRNA is the presence of 5’ triphosphates on the transcribed dsRNA species instead of the 5’ OH present in synthesized 50-bp dsRNA. In support of the impact of 5’ phosphates on transport and/or silencing, addition of 5’ monophosphates to synthesized 50-bp dsRNA injected into parental circulation reduced the dependence on RME-2 for silencing in progeny (Extended Data Fig. 4e,f), potentially by enhancing the ability of synthesized dsRNA to be imported by other dsRNA importers (e.g. SID-1) in the absence of RME-2. Thus, the requirements for dsRNA entry into the germline and subsequent silencing vary for different lengths and/or chemical forms of dsRNA (see summary in Extended Data Table 1).

Fluorescently labeled 50-bp dsRNA delivered into parental circulation localized within intestinal cells in progeny (Fig. 3c, top left), as has been observed for vitellogenin proteins39 and fluorescent dyes40. Accumulation of fluorescently labeled dsRNA was also detected at the apical membrane of the intestine, which could reflect exocytosis of dsRNA into the lumen of developing intestinal cells. However, separation of the fluorescent label from dsRNA catalyzed by cellular enzymes cannot be excluded. Therefore, to dissect differences, if any, between the transport of unlabeled short dsRNA (synthesized 50-bp with 5’OH) and mixed dsRNA (mixture transcribed in vitro using ∼1 kb DNA template) we injected unc-22-dsRNA into animals with mutations in genes that play roles in the import of dsRNA. We found that maternal SID-1 was required for silencing by short dsRNA in progeny (Fig. 3c, bottom, left bars), suggesting that the SID-1-dependent entry of short dsRNA into the cytosol likely occurs in the injected parent or during early development in progeny. Uptake of dsRNA from the intestinal lumen requires SID-2, a transmembrane protein located in the apical membranes of intestinal cells17,18. We found that SID-2 was not required for most silencing in progeny by short or mixed dsRNA injected into parental circulation (Fig. 3c, top right and bottom). Exit of dsRNA from intracellular vesicles requires SID-5, a transmembrane protein located in endolysosomal membranes41. Silencing in wild-type animals was comparable to silencing in sid-5(-) animals (Fig. 3c, top right). However, when animals that lacked SID-1 were injected, SID-5 was required in progeny for silencing by mixed dsRNA from parental circulation (Fig. 3c, bottom, right bars; as also reported in ref.29). Since dsRNA is expected to be present in vesicles upon entry through RME-2 in the absence of SID-128,29, this observation suggests that SID-5 is required for the release of mixed dsRNA from inherited vesicles in progeny.

In summary, injected extracellular dsRNA can enter the germline in parents and be transmitted to progeny through two routes with different substrate selectivity. One route is preferentially used by short dsRNA and relies on RME-2-mediated endocytosis of dsRNA into oocytes, where early exit from vesicles is required for silencing in progeny as evidenced by the need for maternal SID-1 (Fig. 3d, blue). The other route appears to exclude short dsRNA but allows mixed dsRNA entry into the cytosol in the parental germline through SID-1 and exit from inherited vesicles in progeny through a process that requires both zygotic SID-1 and SID-5 (Fig. 3d, grey; ref.29).

Expression of SID-1 is consistent with a role in the intergenerational transport of extracellular dsRNA

All routes of dsRNA transport deduced using the experimental addition of dsRNA ultimately require SID-1 for entry into the cytosol. The proposed model (Fig. 3d) for dsRNA transport into the germline and to progeny suggests the expression pattern of SID-1 is likely to include the germline. We used Cas9-mediated genome editing to insert a piRNA-resistant mCherry sequence42,43 into the sid-1 coding sequence (Fig. 4a) to observe the endogenous expression pattern of SID-1::mCherry. This fusion protein was detectably functional in contrast to other attempts at tagging SID-1 (see Technical comments in Methods). Fluorescence from SID-1::mCherry was visible in the proximal and distal regions of the adult germline (Fig. 4b). Expression also progressively increased during development with tissue-specific enrichment in the developing embryo (Fig. 4c, left), becoming ubiquitous in hatched L1 larvae (Fig. 4c, middle,d). SID-1::mCherry was not easily detectable in the germline during later larval development (Fig. 4c, middle and right). In combination with the expression of RME-2 observed in the proximal germline (Extended Data Fig. 1d; ref32), this expression pattern of SID-1 is consistent with the entry of most dsRNA from circulation of adult animals into the germline, followed by activity of transport mechanisms in developing embryos and early larvae that inherit parental dsRNA.

The expression pattern of SID-1 varies during development.

a, Schematic depicting insertion of mCherry sequence that lacks piRNA binding sites42,43 (jam195[mCherryΔpi]) into the sid-1 gene using Cas9-mediated genome editing. b and c, Representative images showing fluorescence from SID-1::mCherry (black) in (b) the adult gonad arm, (c, left) developing embryos, (c, middle) L1-stage animals, or (c, right) L4-stage animals with sid-1(jam195[mCherryΔpi]) compared to autofluorescence in wild-type animals of the same stages. Numbers (n) of each stage imaged are indicated (100% of animals exhibited the depicted expression patterns). For animals imaged in b, the distal germline was obstructed by the intestine in 1/10 sid-1(jam195[mCherryΔpi]) and 5/9 wild-type animals. d, Airyscan image of an L1-staged animal assembled by stitching four different Z-stacks after depth-coding and taking maximum projections, illustrating the expression of SID-1::mCherry throughout the animal. Scale bar for adult gonad arms in b and embryos in c, 20 μm; scale bar for larvae in c, 50 μm andin d, 10 μm. Also see Technical comments on “Making a sid-1 translational reporter” in Methods.

To determine if acute, induced expression rather than developmental expression of SID-1 can be sufficient for the import of dsRNA into the germline, we engineered the endogenous sid-1 gene to transcribe a fusion transcript with an aptamer-regulated ribozyme (Extended Data Fig. 6a, left) that cleaves itself when not bound to tetracycline (Extended Data Fig. 6a, right) (based on ref.44). Exposing these animals to tetracycline enabled silencing by dsRNA in somatic tissues (hypodermis: Extended Data Fig. 6b, left; body-wall muscles: Extended Data Fig. 6b, right), indicative of stabilization of sid-1 mRNA, production of SID-1 protein, and subsequent dsRNA import in somatic cells. However, such tetracycline-induced silencing was not detectable in the germline (Extended Data Fig. 6c-f). Yet, similar tagging of the ubiquitously expressed gene gtbp-1::gfp resulted in detectable rescue of expression within the germline by tetracycline (Extended Data Fig. 6g). A possible explanation for the poor rescue of SID-1 activity within the germline is that post-transcriptional mechanisms targeting sid-1 mRNA in the germline but not in somatic cells interfere with tetracycline-dependent stabilization of the sid-1 transcript (e.g., piRNA-based regulation of sid-1 mRNA45,46), or that acute stabilization of the sid-1 transcript does not override developmental regulation of SID-1 translation.

Additional attempts to tag the SID-1 protein guided by structure and to modulate sid-1 transcripts guided by post-transcriptional regulatory interactions could improve control of dsRNA transport between cells. Nevertheless, the developmentally regulated expression observed for both SID-1 and RME-2 in the germline is consistent with intergenerational or transgenerational effects of endogenous dsRNA from parental circulation after development of the adult germline.

Temporary loss of sid-1 causes a transgenerational increase in the levels of mRNA from two germline genes

To understand how the dsRNA importer SID-1 might be used in endogenous gene regulation across generations, we searched for sid-1-dependent changes in gene expression that could be heritable (Fig. 5, Extended Data Fig. 3 and Extended Data Fig. 7). To control for genetic background (see Technical comments in Methods), we used Cas9-mediated genome editing to delete the entire sid-1 coding sequence or introduce a nonsense mutation in cohorts of the same wild-type animals. By comparing polyA+ RNA from this wild type with that of the newly generated sid-1(jam113[deletion]) (Fig. 5a,b and Extended Data Fig. 7a) or sid-1(jam80[nonsense]) (Fig. 5a-c) animals, we found that 26 genes were significantly (q < 0.05) misregulated in sid-1(jam113[deletion]) (Extended Data Fig. 7b) and 6 in sid-1(jam80[nonsense]) (Fig. 5d, top), both including sid-1 (a list of significantly altered genes is in Extended Data Table 2). The most upregulated gene in sid-1(jam113[deletion]), F14F9.5 (Extended Data Fig. 7b), which is located immediately 3’ to sid-1 in the genome, was only misregulated in the deletion mutant and not in the nonsense mutant (Extended Data Fig. 7c, left). Both mutants, however, were equally defective for silencing by ingested dsRNA (Fig. 5b). This observation suggests that while both mutations result in loss of SID-1 protein, the deletion of sid-1 also changes local regulatory sequences (potentially explaining upregulation of the neighboring gene F14F9.5) and eliminates sid-1 mRNA, which could participate in RNA-based regulatory interactions within the germline45,46. Nevertheless, we could detect two genes that were upregulated in both sid-1(jam113[deletion]) and sid-1(jam80[nonsense]) animals (red in Fig. 5d, top, and Extended Data Fig. 7b): the identical loci W09B7.2/F07B7.2 (Extended Data Fig. 7c, middle), and Y102A5C.36 (Extended Data Fig. 7c, right). Intriguingly, another gene cls-3 also changed in both mutants (Extended Data Table 2) but in different directions (∼3.4-fold decrease in the sid-1(jam80[nonsense]) mutant but a ∼5.8-fold increase in the sid-1(jam113[deletion]) mutant), suggesting that the direction of change in expression can vary. Conservatively, we began by analyzing only the two genes with mRNA levels that changed in the same direction in both sid-1 mutants. Both W09B7.2/F07B7.2 and Y102A5C.36 have been reported47 to be expressed within the germline (Extended Data Fig. 7d, left) and regulated by endogenous small RNAs (Extended Data Fig. 7d, middle and right). Spliced mRNA levels measured at a later generation using RT-qPCR demonstrated that both transcripts were upregulated in sid-1(jam80[nonsense]) animals compared to wild-type animals as expected (Fig. 5e), but no upregulation was detectable in sid-1(jam113[deletion]) animals (Extended Data Fig. 7e). This difference between the two sid-1 mutants could reflect increased variation in expression (as was observed for cls-3 using RNAseq) or could reflect complex effects caused by deletion of sid-1 DNA (e.g., F14F9.5 overexpression, loss of sid-1 mRNA, etc.) that could be independent of SID-1 protein function.

Ancestral loss of SID-1 causes transgenerational changes in the mRNA levels of two germline genes that are subject to RNA regulation.

a, Schematic of modifications at the sid-1 gene generated using Cas9-mediated genome editing. Deletion of the entire coding sequence (jam113[deletion]), a nonsense mutation (jam80[nonsense]), and its reversion to wild-type sequence (jam86[revertant]) are depicted. b, Fractions of animals with the indicated genotypes that show silencing in response to unc-22-dsRNA (grey) or bli-1-dsRNA (black). Numbers of animals scored (n), significant differences using two-tailed test with Wilson’s estimates for single proportions (asterisks, P < 0.05 with Bonferroni correction) and 95% CI (error bars) are indicated. c, Principal components explaining the variance between wild type (black), sid-1(jam80[nonsense]) (red), and sid-1(jam86[revertant]) (grey) polyA+ RNA samples. Almost all of the variance between samples is explained by PC 1. d, Volcano plots of changes in the abundance of polyA+ RNA in sid-1(jam80[nonsense]) (top) and sid-1(jam86[revertant]) (bottom) animals compared with wild-type animals (black, q < 0.05; red, both q < 0.05 and change in the same direction in sid-1(jam80[nonsense]) and sid-1(jam113[deletion]); see Extended Data Fig. 7). While sid-1 transcript levels in sid-1(jam86[revertant]) are comparable to that in wild type (grey), sdg-1 (W09B7.2/F07B7.2) and sdg-2 (Y102A5C.36) transcript levels remain elevated in sid-1(jam86[revertant]) (red). e, Levels of spliced sid-1 (top), sdg-1 (middle) and sdg-2 (bottom) transcripts measured using RT-qPCR. The median of three technical replicates is plotted for each of three biological replicates (bar indicates median) assayed before and after 1 year of passaging animals (year 1, dark grey; year 2, light grey). Asterisks indicate P < 0.05 with Bonferroni correction using two-tailed Student’s t-test. f, Heatmap showing changes in the levels of transcripts (total RNA or mRNA) or antisense small RNAs (22G RNA) from sid-1, sdg-1, sdg-2, and tbb-2 (abundant germline transcript for comparison). Fold changes (expressed as LogFC, indicating log2 for (m)RNA, log10 for piRNA binding, and log10 for 22G RNA) were deduced by integrating reports (study) of 21 experiments that identify subsets of genes as being subject to RNA-mediated regulation within the germline (# genes). These prior studies include comparisons of RNA or 22G RNA from wild-type animals with that from mutant animals (e.g., mut-16(-) 22G RNA), biochemical detection of piRNA binding to transcripts (piRNA-bound mRNA), and biochemical detection of 22G RNA binding to an Argonaute (HRDE-1-bound 22G RNA). ‘NS’ indicates cases where changes, if any, were not significant based on the criteria used in the study. A conservative value of 2-fold is assigned to all genes reported as changing >2-fold in ref.58.

To determine if changes in W09B7.2/F07B7.2 and Y102A5C.36 expression were heritable, we reverted the sid-1 nonsense mutation to wild-type sequence using Cas9-mediated genome editing. This immediately restored silencing by ingested dsRNA (Fig. 5b) with concomitant recovery of sid-1 mRNA to wild-type levels (Fig. 5e, top). In contrast, changes in both W09B7.2/F07B7.2 and Y102A5C.36 expression persisted (Fig. 5d, bottom) even after one year of passaging the reverted animals (sid-1(jam86[revertant])) (i.e., after >100 generations, Fig. 5e, middle and bottom). Thus, the sid-1-dependent accumulation of mRNA from these two germline genes persisted for many generations, likely through mechanisms that maintain heritable epigenetic changes. We hereafter refer to these sid-1-dependent genes (sdg) that show heritable epigenetic changes in response to temporary loss of SID-1 as sdg-1 (W09B7.2/F07B7.2) and sdg-2 (Y102A5C.36).

The sid-1-dependent gene sdg-1 is affected by many factors that regulate RNA silencing in the germline

To determine if expression of sdg-1 and sdg-2 is regulated by other proteins that play a role in RNA silencing within the germline, we examined 21 published datasets30,46,48-58 that reported changes that depend on such proteins. For each dataset, we determined if the lists of genes reported as showing significant changes in mutants compared to the respective wild types included sdg-1 and/or sdg-2. This analysis revealed that changes in mRNA and/or antisense small RNAs of sdg-1 were detected in 20 of the 21 datasets while changes in sdg-2 were observed in 9 of 21 (Fig. 5f). When detected, changes in sdg-2 were in the same direction as changes in sdg-1, suggestive of similar regulation of both genes.

RNAs transcribed in the germline can be recognized as they exit the nuclear pores by piRNAs bound to the Argonaute PRG-1, which recruits them for regulation by antisense small RNAs called 22G RNA made by proteins organized within perinuclear germ granules (reviewed in ref.59). Interaction with piRNAs was detected for RNA from sid-1, sdg-1, and sdg-2, and the control gene tbb-2 using crosslinking, ligation, and sequencing of hybrids53 (Fig. 5f), consistent with their germline expression. Depletion of downstream 22G RNAs targeting both sid-1 and sdg-1 was detectable upon loss of the germ granule component MUT-1651 (Fig. 5f). Both genes were among the top 500 genes targeted by 22G RNAs bound by the secondary Argonaute HRDE-1/WAGO-930 (Fig. 5f), suggesting similar downregulation of both genes using 22G RNAs. Furthermore, multiple datasets support downregulation of sdg-1 within the germline by HRDE-1/WAGO-9-bound 22G RNAs in the absence of PRG-1. One, loss of HRDE-1/WAGO-9 increased sdg-1 RNA in whole animals58 (Fig. 5f) and in dissected gonads57 (Fig. 5f). Two, loss of PRG-1 decreased sdg-1 RNA (Fig. 5f) and increased 22G RNAs that are antisense to sdg-1 (Fig. 5f) in dissected gonads52. Three, although animals that lack PRG-1 become progressively sterile, the increase in sdg-1 22G RNA persisted in near-sterile animals (Fig. 5f, near-sterile in ref.52), and this increase was eliminated upon additional loss of HRDE-1/WAGO-9 (Fig. 5f, near-sterile in ref.52).

As expected for sid-1-dependent downregulation of sdg-1, multiple datasets reveal an inverse relationship between the two genes. In animals lacking PRG-1, sid-1 RNA levels increased and sid-1 22G RNAs decreased55 (Fig. 5f), but both sdg-1 RNA and sdg-2 RNA levels decreased along with an increase in 22G RNAs51-56 (Fig. 5f). This inverse relationship between sid-1 and sdg-1 RNA regulation is also observed when many components of germ granules are mutated as indicated by changes in 22G RNA upon loss of the embryonic P granule components MEG-3/-446 (Fig. 5f), the PRG-1 interactor DEPS-151 (Fig. 5f), or the Z granule component ZSP-150 (Fig. 5f; also known as PID-260).

In addition to the above studies, pioneering studies that used microarrays identified sdg-1 as upregulated in animals lacking the germ granule component DEPS-149 (Fig. 5f) and in animals lacking the dsRNA-binding protein RDE-4 (Fig. 5f; second-most upregulated in ref.48), which recruits dsRNA imported through SID-1 and other intracellular dsRNA for processing and eventual gene silencing. Animals that lack RDE-4 show a ∼47.5-fold increase in sdg-1 RNA48. A reduction in RDE-4 activity could also contribute to the ∼11.6-fold increase in sdg-1 RNA seen in deps-1(-) animals because these animals also show a ∼3.2-fold decrease in rde-4 RNA49 (one of 13 downregulated genes). These observations support the idea that appropriate regulation of sdg-1 RNA requires both piRNA-mediated processes that act via germ granule components such as DEPS-1 and dsRNA-mediated processes that use SID-1 and RDE-4.

In summary, the levels of sdg-1 RNA are detectably regulated by the dsRNA-selective importer SID-1, the dsRNA-binding protein RDE-4, and the piRNA-binding Argonaute PRG-1. Presence of dsRNA-mediated regulation or loss of piRNA-mediated regulation enhances MUT-16-dependent production of secondary small RNAs that bind the secondary Argonaute HRDE-1/WAGO-9. Consistent with downregulation of these sid-1-dependent transcripts by SID-1, disruption of many components of germ granules results in opposite effects on these transcripts and sid-1 RNA. Intriguingly, a search of protein interaction studies revealed that the SDG-1 protein is among the interactors of two germ granule components: PID-2 by immunoprecipitation60 (also known as ZSP-150) and DEPS-1 by proximity labeling61. Thus, one possibility suggested by these observations is that reduction of sdg-1 RNA via SID-1 alters the amount of SDG-1 protein, which could interact with components of germ granules to mediate RNA regulation within the germline of wild-type animals.

Regulation of sdg-1 RNA is susceptible to epigenetic changes that last for many generations

SDG-1 is encoded by a gene located within a retrotransposon (Extended Data Fig. 8a) that is within a duplicated ∼40 kb region and has two recognizable paralogs (Extended Data Fig. 8b). To facilitate analysis of SDG-1 expression, we tagged both loci that encode SDG-1 with mCherry coding sequences lacking piRNA-binding sites42,43 (mCherryΔpi) (Extended Data Fig. 8c,d). This tagging resulted in the expression of sdg-1::mCherryΔpi mRNA being ∼16-fold higher than sdg-1 mRNA (Extended Data Fig. 8e), potentially because of the reduction in the overall density of piRNA-binding sites per transcript, the additional introns included in mCherryΔpi (based on refs.62,63), and/or other unknown factors. Fluorescence from SDG-1::mCherry was observed in the germline of adult animals (Fig. 6a). However, animals showed variation in SDG-1::mCherry expression between their two gonad arms (Fig. 6a, middle shows bright anterior (20% of animals) and right shows bright posterior (6% of animals)). A contributing feature for the observed stochasticity could be the location of sdg-1 within a duplicated region (Extended Data Fig. 8a), as suggested by similar stochastic RNA silencing of multi-copy transgenes but not single-copy transgenes64. Despite this variation, unbiased passaging of self-progeny for more than 18 generations continuously preserved SDG-1::mCherry expression in an otherwise wild-type background (Fig. 6b). In contrast, mating, which can perturb RNA regulation within the germline in cross progeny43, caused dramatic changes in sdg-1 expression that persisted in descendants (Fig. 6c). Mating animals that express SDG-1::mCherry with wild-type animals resulted in heritable changes along lineages that express sdg-1::mCherryΔpi mRNA or that express sdg-1 mRNA (Fig. 6c and Extended Data Fig. 9a). This discovery of mating-induced perturbation in gene expression raises caution in interpreting past studies (summarized in Fig. 5f) where changes in multiple independent isolates were not examined. Nevertheless, when we used genetic crosses to determine the impact of mutations on sdg-1 expression, we observed reduced SDG-1::mCherry fluorescence in mutants predicted to have reduced levels of intracellular dsRNA (Fig. 6d, sid-2(-), sid-5(-)) or reduced processing of intracellular dsRNA (Fig. 6d, eri-1(-)). In contrast, we observed an increase in SDG-1::mCherry fluorescence in animals lacking MUT-16 (Fig. 6d). Finally, animals lacking RME-2, which lack the ability to import many maternal factors (e.g., lipids, proteins, RNAs, etc.), also showed an increase in SDG-1::mCherry fluorescence (Fig. 6d).

The sdg-1 gene is prone to stochastic changes in gene expression that can become heritable.

a, Representative images showing fluorescence of SDG-1::mCherry (black) in a wild-type background. While most animals showed symmetric expression in the germline (left), animals with >2-fold difference in fluorescence between both gonad arms (bright anterior, middle and bright posterior, right) were also observed. Punctate fluorescence in the intestine likely represents autofluorescence. Scale bar, 50 μm. b, Quantification of SDG-1::mCherry fluorescence intensity (arbitrary units, a.u.) in adult gonad arms (anterior arm, dark grey; posterior arm, light grey) of sdg-1(jam137[mCherryΔpi]) animals starting in one generation (x) and continuing in successive generations as indicated. Numbers of gonad arms quantified (n) is indicated. Expression in one generation was not significantly different when compared to that in the previous tested generation using Mann-Whitney U test for two-sided comparisons and Bonferroni correction. c, Lineages and estimated relative sdg-1 expression 10 generations after mating wild-type (open circle) males with sdg-1::mCherryΔpi (filled circle) hermaphrodites and vice versa, and isolating sdg-1(+) or sdg-1::mCherry animals from F1 heterozygotes (half-filled circle). Expression of sdg-1 in the F10 generation was measured by RT-qPCR of sdg-1 mRNA purified from pooled wild-type animals of mixed stages or by quantification of SDG-1::mCherry fluorescence in gonad arms of adult sdg-1::mCherryΔpi animals. Relative levels of sdg-1 mRNA and SDG-1::mCherry fluorescence intensity were converted to units of estimated relative sdg-1 expression (see Methods) for comparison. See Extended Data Fig. 9a for raw data. d-f, Fluorescence intensity measurements (quantified as in b) in adult animals with sdg-1::mCherryΔpi (+) and additionally with mutations in genes introduced through genetic crosses (in regulators of dsRNA import rme-2, sid-2 or sid-5, or in regulators of RNA silencing mut-16 or eri-1) or through genome editing (in regulators of dsRNA import sid-1 or sid-3, or in regulators of RNA silencing rde-1 or deps-1). Asterisks indicate P < 0.05 with Bonferroni correction using Mann-Whitney U test for two-sided comparisons between animals with sdg-1::mCherryΔpi (+) and animals with additional mutations. Nonsense mutations (nonsense) or deletions (deletion) introduced through genetic crosses (isolate numbers #1, #2, etc. in d) or genome editing (different alleles in e and f) and numbers of gonad arms (n) quantified for each isolate are indicated. Mutations in genes required for dsRNA import or subsequent silencing resulted in fewer animals with asymmetric fluorescence between gonad arms (a combined proportion of 21/197 for sid-1, sid-3, rde-1 and deps-1 mutants versus 22/84 for wild type, P = 0.0009 using two-tailed test with Wilson’s estimates for single proportions). Animals with at least one gonad arm brighter than the dimmest wild-type gonad arm in a and with asymmetric gonad arms were found in different genotypes (anterior bright: sid-1(-) – 5/122, sid-3(-) – 1/29, rde-1(-) – 2/22, deps-1(-) – 4/24, and posterior bright: sid-1(-) – 6/122, rde-1(-) – 2/22, deps-1(-) – 1/24). g, Fluorescence intensity measurements as in b of animals with sdg-1::mCherryΔpi that show loss of fluorescence when a nonsense mutation is introduced in sid-1 using genome editing ∼30 generations (gen.) later remain changed despite reversion of sid-1 nonsense mutation to wild-type sequence after ∼20 additional generations. Subsequent mutation of deps-1 after another ∼110 generations restored SDG-1::mCherry fluorescence to wild-type levels. Also see Extended Data Fig. 8 and Extended Data Fig. 9.

To avoid mating-induced perturbations of RNA regulation within the germline, we used Cas9-mediated genome editing to introduce mutations into animals that express SDG-1::mCherry in an otherwise wild-type background. Use of this approach to mutate a control gene with no known roles in RNA regulation within the germline resulted in similar levels of SDG-1::mCherry fluorescence in multiple isolates of animals with and without the mutation (Extended Data Fig. 9b). In contrast, mutating sid-1 using Cas9-mediated genome editing caused a range of expression levels in different isolates when compared with sid-1(+) animals (Fig. 6e - 6 isolates lower, 2 isolates comparable, and 1 isolate higher), which differs from the increase in sdg-1 mRNA observed upon SID-1 loss in the single isolate examined earlier (Fig. 5). One explanation for these observations is that in the absence of SID-1, the levels of SDG-1 vary in different isolates because the sdg-1 gene becomes prone to heritable epigenetic changes, resulting in either high or low expression states being stabilized in different isolates. Alternatively, it is possible that the ∼16-fold overexpression of sdg-1::mCherryΔpi mRNA perturbs RNA-mediated epigenetic regulation, potentially indicative of a role for the SDG-1 protein in such regulation. Mutating sid-3 also lowered the levels of SDG-1::mCherry in one isolate, but caused no detectable change in another (Fig. 6f). While both isolates with loss of RDE-1 showed lower levels of SDG-1::mCherry, both isolates with loss of the germ granule component DEPS-1 showed higher levels of SDG-1::mCherry (Fig. 6f). These experiments also reveal that in most isolates of animals expected to have reduced levels of intracellular dsRNA (sid-3(-), sid-1(-)) or reduced processing of intracellular dsRNA (rde-1(-)), SDG-1::mCherry fluorescence is reduced. In contrast, isolates lacking DEPS-1 showed increased expression of SDG-1::mCherry.

Collectively, the observations on SDG-1 expression using mutants suggest that the uptake and processing of intracellular dsRNA (mediated by SID-1, SID-2, SID-3, SID-5, RDE-1, and ERI-1) and the function of germ granules (mediated by MUT-16 and DEPS-1) are both necessary to maintain intermediate levels of SDG-1 expression across generations. Once the levels of SDG-1::mCherry were reduced upon loss of SID-1, downregulation persisted across generations even after restoration of wild-type SID-1 (Fig. 6g), just as the upregulation of untagged sdg-1 mRNA also persisted (Fig. 5). Despite >100 generations of such persistent silencing, the expression of SDG-1::mCherry could be restored by mutating deps-1 (Fig. 6g), implicating small RNA-based regulation and germ granules in the maintenance of new epigenetic states established upon loss of SID-1.

SID-1-dependent genes, including SDG-1, could alter RNA-mediated regulation in the germline

Since SDG-1 interacts with regulators of RNA silencing (ZSP-1/PID-260 and DEPS-161), loss of SID-1 could both block the entry of extracellular dsRNA into the cytosol and change intracellular RNA regulation through SID-1-dependent genes such as SDG-1. In support of this possibility, intracellular delivery of dsRNA through injection into the syncytial germline of sid-1(-) animals showed a weak defect in silencing (Fig. 2D in ref.28 and Fig. 1 in ref.29). To examine if changes in the levels of SDG-1 alone could account for such a defect in silencing by intracellular dsRNA in the germline, we exposed independently-generated sdg-1 deletion animals and animals that overexpress sdg-1 (i.e., with sdg-1::mCherryΔpi) to dsRNA matching the germline gene pos-1 for ∼16 hours. This short exposure to pos-1-dsRNA caused intermediate levels of silencing in wild-type animals and comparable intermediate silencing in sdg-1 mutant and sdg-1 overexpressing animals (Fig. 7a), suggesting that changes in sdg-1 levels alone are not sufficient to cause a defect in silencing by dsRNA that is detectable using this assay. Alternatively, the previously described defect in silencing by intracellular dsRNA in sid-1(-) animals28,29 could be through the promotion of competing piRNA-mediated gene regulation in the absence of SID-1. This notion that dsRNA-mediated and piRNA-mediated gene regulation compete in the germline is supported by a study that demonstrated that loss of PRG-1 enhances heritable RNA silencing by dsRNA65. To test if loss of SID-1 or a sid-1-dependent gene enhances piRNA-mediated silencing, we used an experimental system which initiates piRNA-mediated silencing of the two-gene operon described in Fig. 2 through mating, independent of externally provided dsRNA43. We found that sid-1(-) animals exhibited enhanced mating-induced silencing (Fig. 7b, top right: ∼50% off in sid-1(+) vs 100% off in sid-1(qt9[nonsense])), while animals lacking sdg-1 showed a small reduction in mating-induced silencing that was not statistically significant (Fig. 7b, bottom right, ∼40% off in sdg-1(+) vs ∼30% off in sdg-1(jam232[deletion])). Taken together, these results support the model that loss of SID-1 weakly inhibits silencing by intracellular dsRNA28,29 but enhances silencing by piRNAs within the germline, potentially through the action of multiple altered SID-1-dependent genes that collectively promote piRNA-mediated gene regulation.

SID-1 modifies RNA regulation within the germline, potentially through sdg-1 and other sid-1-dependent genes.

a, (left) Schematic of assay for sensitive detection of pos-1 silencing by ingested dsRNA. (right) Numbers of developed progeny (> 3rd larval stage) laid by wild-type animals, animals with a deletion (Δ) in sdg-1 (jam232, jam241, jam242) or animals with overexpression (sdg-1::mCherryΔpi) of sdg-1 exposed to pos-1 dsRNA (red) or control dsRNA (black) for 16 hours are plotted. Asterisks indicate P < 0.05 using Mann-Whitney U test for two-sided comparisons with Bonferroni correction. b, Cross progeny males that inherited the mex-5p::mCherry::h2b::gfp::h2b transgene (T)43 (also used in Fig. 2) from maternal (left) or paternal (right) parents, both of wild-type, sid-1(-), or sdg-1(-) background, were scored for expression of mCherry and GFP (bright, dim, off). Wild-type data for top set (n = 77 and n = 33) are replotted from ref.43 for comparison. Dashed line separates independent experiments. Asterisk indicates P < 0.05 with Bonferroni correction using χ2 test; n.s. indicates not significant. c, Representative AiryScan images of the distal germline (left; scale bar, 10 µm) or single germline nuclei (right; scale bar, 2 µm) showing SDG-1::mCherry alone (top) or with GFP::ZNFX-1 (bottom, merge and single channel images). The number of animals imaged (n) and the percentage that show enrichment of SDG-1::mCherry in perinuclear foci are indicated. Sites of SDG-1::mCherry enrichment coincide with GFP::ZNFX-1 localization. Boxes in left mark the nuclei shown in right. d, Representative images showing entry of SDG-1::mCherry into the nucleus in -1 oocytes (left) and upon pronuclear fusion in early embryos during the time course indicated (right). Numbers of germlines and embryos imaged are indicated. Scale bars, 20 µm. Also see Extended Data Movies 1-4. e, Representative image of the hermaphrodite germline in animals with a translational (left) or transcriptional (right) reporter of sdg-1. Scale bars, 20 µm. Apparent extracellular punctae of SDG-1::mCherry and mCherry surrounding the proximal germline requires further study, but could be non-specific because similar localization is observed in animals with other promoters driving mCherry expression, but not GFP expression, in the germline (data not shown). The numbers of animals with the depicted fluorescence pattern are indicated. f and g, Response of the transcriptional sdg-1 reporter (sdg-1p::mCherryΔpi[sdg-1(Δ)]::sdg-1 3’ UTR) to the addition of unc-22-dsRNA (f) or loss of rde-4 (g). Quantification and asterisk are as in Fig. 6. h, Models for dsRNA import into the germline (top) and subsequent RNA-mediated regulation of sdg-1 (bottom). See text for details.

RNA regulation within the germline using piRNAs relies on phase-separated granules (reviewed in ref.66). To determine if the identification of SDG-1 as an interactor of the Z-granule component PID-260/ZSP-150 and potentially of the P-granule-adjacent protein DEPS-161 could be seen as colocalization with Z-granules in vivo, we examined the localization of SDG-1::mCherry within the cytoplasm at high resolution using AiryScan imaging67. SDG-1::mCherry was enriched in perinuclear foci in many animals (Fig. 7c, top; 7 of 9 animals) and these sites of enrichment colocalized with the Z-granule marker GFP::ZNFX-168,69 (Fig. 7c, bottom; 100% colocalized in 10 of 12 animals with enrichment). Time-course imaging revealed re-localization of SDG-1::mCherry into the nucleus from the cytoplasm in the -1 oocyte, which will be the first to be fertilized (Fig. 7d, left and Extended Data Movie 1), along with subsequent exclusion from the nucleus before the maternal and paternal pronuclei meet (Fig. 7d, right and Extended Data Movie 2). Such dynamic entry into the nucleus followed by exclusion from the nucleus also occurred during early cell divisions in the developing embryo (Extended Data Movies 3,4). The timing of nuclear entry of SDG-1::mCherry coincides with the nuclear envelope breakdown events that occur during fertilization and early development70. The sdg-1 coding sequence was required for regulated nuclear entry because deletion of the sdg-1 open reading frame in sdg-1::mCherryΔpi animals resulted in mCherry expression throughout the germline in both the cytoplasm and nuclei (Fig. 7e). Nuclear localization of the SDG-1 protein in the -1 oocyte is like that of the essential Argonaute CSR-1b71, thought to play a role in protecting transcripts from silencing. Exposure to ingested dsRNA did not alter the expression of sdg-1p::mCherryΔpi[sdg-1(Δ)]::sdg-1 3’ UTR (Fig. 7f) but loss of rde-4 perturbed expression such that one isolate showed loss of expression while the other showed enhanced expression (Fig. 7g). These results suggest that while the sdg-1 open reading frame is required for exclusion from germline nuclei, it is not required for the response of the sdg-1 gene to changes in intracellular dsRNA. Together, these observations on the levels and localizations of SDG-1 raise the possibility that this protein is actively regulated by extracellular and intracellular dsRNA with a role in heritable RNA silencing, potentially with Z granules, and additional cell-cycle coupled roles during early development, potentially through nucleocytoplasmic shuttling.

Discussion

Our analyses suggest a model for the impact of dsRNA from parental circulation on descendants (Fig. 7h). Extracellular dsRNA can accumulate in parental circulation through regulated secretion from neurons (e.g., oxidative damage promotes accumulation) and potentially other tissues. Uptake into the germline is RME-2-dependent for some forms of dsRNA (blue in Fig. 7h), but RME-2-independent for other forms of dsRNA (grey in Fig. 7h). While all forms of dsRNA require SID-1 for the entry of dsRNA into the cytosol in progeny, RME-2-dependent dsRNA also require SID-5. Such dsRNA from parental circulation along with other intracellular dsRNA are processed with the help of the dsRNA-binding protein RDE-4 in progeny (Fig. 7h, bottom). This processing of dsRNA regulates the SID-1-dependent gene sdg-1 by reducing variation between animals in the levels of sdg-1 mRNA, although the dsRNA sequences likely need not match sdg-1 because no such dsRNA was reported in the published dsRNAome72 and because this regulation can occur in the absence of the sdg-1 open reading frame. The sdg-1 gene is located within a retrotransposon that is also targeted by heritable RNA silencing, but the SDG-1 protein colocalizes with perinuclear granules called Z granules, which are required for heritable RNA silencing. Consistent with the SDG-1 protein promoting heritable RNA silencing, the sdg-1::mCherryΔpi gene is highly susceptible to mating-induced silencing, potentially owing to the ∼16-fold higher levels of SDG-1::mCherry compared with that of SDG-1 in wild-type animals. In agreement with this proposal, SID-1 limits heritable RNA silencing because loss of SID-1 enhances mating-induced silencing (Fig. 7b, top). Since the sdg-1 gene is located within a retrotransposon that is targeted by heritable RNA silencing, this mechanism for regulating the regulators of heritable RNA silencing such as SDG-1 (and potentially other SDGs) reveals a strategy for tuning an autoregulatory loop for heritable RNA silencing by using competing dsRNA processing. Intriguingly, SDG-1 becomes enriched within nuclei upon nuclear envelope breakdown during fertilization and during early cell divisions in embryos (Fig. 7h, bottom right) with active exclusion from nuclei after each reformation of the nuclear envelope, suggestive of additional roles for this retrotransposon-encoded protein.

While the physiological conditions that promote secretion of dsRNA are not known, the discovery that oxidative damage of neurons can enhance the secretion of dsRNA suggests that disruption of cell structures by oxidative damage (e.g., membrane integrity) or initiation of cellular processes that repair oxidative damage (e.g., through ejection of damaged macromolecules73) also promote the release of dsRNA. Pathologies of the central nervous system in humans, including cancer, stroke, multiple sclerosis, neurodegenerative disease, and brain injury, have been associated with extracellular RNAs detected in circulation (reviewed in ref.74), although their origins and regulatory consequences, if any, remain unknown. The gene regulatory effects of neuronal dsRNA released upon oxidative damage of neurons provide convenient readouts that can be analyzed to understand neuronal damage and its consequences in animals.

The trafficking of extracellular dsRNA from parent to progeny has spatial specificity, as evidenced by more silencing within the proximal germline (Fig. 2), temporal specificity, as evidenced by the need for dsRNA beyond the fourth larval stage28,29 (Fig. 1), and substrate specificity, as evidenced by the differential requirements for 50-bp dsRNA with 5’ OH versus a mix of longer dsRNAs with 5’ triphosphates (Fig. 3). One possible explanation for these constraints could be that proteins mediating dsRNA transport or subsequent silencing differ in their availability during development and in their affinities for different substrates. For example, SID-1, which was not detected in the developing larval germline but was detected in the adult germline (Fig. 4), has an extracellular domain that binds dsRNA75 and could prefer dsRNA molecules with 5’ phosphates and/or long dsRNA. Although the selectivity uncovered here could apply to all dsRNA delivered into the extracellular space of C. elegans from any source (see Extended Data Table 1 for constraints on intergenerational dsRNA transport), the chemistry of the delivered dsRNA could be modified by yet unidentified enzymes in vivo to overcome these requirements. Tracking labeled dsRNA with diverse chemistries from parental circulation to progeny could allow correlation of differences observed in progeny silencing to differences in intergenerational trafficking. The germline is a major site of dsRNA import in C. elegans as evidenced by three key observations: the expression of SID-1 in the germline (Fig. 4), heritable misregulation of germline genes in sid-1(-) animals (Fig. 5 and Fig. 6), and accumulation of fluorescently-labeled dsRNA from the extracellular space in the germline28,29. As a result, sid-1(-) animals could have a defect in the germline that is detectable only under conditions that promote dsRNA transport (e.g., oxidative damage). Multiple physiological defects in the germline and soma of sid-1(-) animals have been reported but have not been widely reproduced, have only been characterized within single generations, or have not been attributed to any specific sid-1-dependent gene(s). These include defects in animals with some misfolded proteins in the endoplasmic reticulum76, in animals exiting the dauer stage77,78, in animals exposed to pathogenic P. aeruginosa79,80,81, in animals exposed to odor82, in intestinal cells that develop in the presence of a multi-copy transgene83, and in animals that overexpress α-synuclein84. RNA-seq experiments in this study and comparisons to those of previous studies suggest that genetic background-dependent changes can obscure genuine sid-1-dependent changes (see Extended Data Fig. 7 and Technical comments in Methods), raising caution in the interpretation of putative sid-1-dependent defects. Comparing multiple sid-1 mutants generated using genome editing with animals in which the mutated sequence has been reverted to wild-type sequence in the same genetic background could provide a firmer basis for the identification of additional sid-1-dependent processes.

Genes expressed within the germline are likely regulated by positive feedback loops that are required to continually produce factors for maintaining germline immortality and for preserving form and function across generations85,86. Thus, germline genes could be particularly vulnerable to heritable epigenetic changes, where deviations in the expression levels of a gene that is regulated by or is part of such feedback loops has the potential to become permanent in descendants. Our analysis of sdg-1 expression suggests that it is part of a regulatory architecture that is susceptible to heritable epigenetic changes through the perturbation of RNA regulation (Fig. 5, Fig. 6, and Fig. 7). Such architectures within the germline could be exploited by ‘selfish’ genetic elements such as retrotransposons to persist across evolution if one of these elements also include genes encoding a regulator. In support of a wider use of such a strategy, a paralog of SDG-1, ZK262.8 (Extended Data Fig. 8b), is also encoded by a gene located within a retrotransposon and its loss along with that of the miRNA-associated Argonaute ALG-2 was reported to be synthetic lethal87. To buffer against heritable epigenetic changes, levels of gene expression would need to be maintained within a particular range for a certain regulatory context. Given the association of SDG-1 protein with germ granule components ZSP-1/PID-2 and DEPS-1, and the maintenance of heritable changes in sdg-1::mCherryΔpi expression by DEPS-1, buffering against changes in gene expression could involve both RNA- and protein-based regulation that tunes the function of perinuclear granules. We therefore speculate that one role for extracellular RNAs that enter germ cells in other systems (e.g., tRNA fragments in mammals5,6,8) could be to similarly buffer against heritable changes in gene expression.

Supporting information

eLife Assessment

In this report, the authors present valuable findings identifying a novel worm-specific protein (sdg-1) that is induced upon loss of dsRNA import via SID-1, but is not required to mediate SID-1 RNA regulatory effects. The genetic and genomic approaches are well-executed. The existing data are solid, but the study would benefit from additional supporting evidence. The manuscript's central findings could also be refined to avoid overstating the results. These findings will be of interest to those working in the germline epigenetic inheritance field.

Reviewer #1 (Public review):

Summary:

In the manuscript "Intergenerational transport of double-stranded RNA limits heritable epigenetic changes" Shugarts and colleagues investigate intergenerational dsRNA transport in the nematode C. elegans. They induce oxidative damage in worms, blocking dsRNA import into cells (and potentially affecting the worms in other ways). Oxidative stress inhibits dsRNA import and the associated heritable regulation of gene expression in the adult germline (Fig. 2). The authors identify a novel gene, sid-1-dependent gene-1 (sdg-1), which is induced upon inhibition of SID-1 (Fig. 3). Both transient inhibition and genetic depletion of SID-1 lead to the upregulation of sdg-1 and a second gene, sdg-2 (Fig. 5). The expression of SDG-1 is variable, potentially indicating buffering regulation. While the expression of Sdg-1 could be consistent with a role in intergenerational transport of dsRNA, neither its overexpression nor loss-of-function impacts dsRNA-mediated silencing (Fig. 7) in the germline. It would be interesting to test if sdg-2 functions redundantly.

In summary, the authors have identified a novel worm-specific protein (sdg-1) that is induced upon loss of dsRNA import via SID-1, but is not required to mediate SID-1 RNA regulatory effects.

Remaining Questions:

• The authors use an experimental system that induces oxidative damage specifically in neurons to release dsRNAs into the circulation. Would the same effect be observed if oxidative damage were induced in other cell types?

• Besides dsRNA, which other RNAs and cellular products (macromolecules and small signalling molecules) are released into the circulation that could affect the observed changes in germ cells?

• SID-1 modifies RNA regulation within the germline (Fig. 7) and upregulates sdg-1 and sdg-2 (Fig. 5). However, SID-1's effects do not appear to be mediated via sdg-1. Testing the role of sdg-2 would be intriguing.

• Are sdg-1 or sdg-2 conserved in other nematodes or potentially in other species? Sdg-1 appears to be encoded or captured by a retro-element in the C. elegans genome and exhibits stochastic expression in different isolates. Is this a recent adaptation in the C. elegans genome, or is it present in other nematodes? Does loss-of-function of sdg-1 or sdg-2 have any observable effect?

Clarification for Readability:

To enhance readability and avoid misunderstandings, it is crucial to specify the model organism and its specific dsRNA pathways that are not conserved in vertebrates:

• In the first sentence of the paragraph "Here, we dissect the intergenerational transport of extracellular dsRNA ...", the authors should specify "in the nematode C. elegans". Unlike vertebrates, which recognise dsRNA as a foreign threat, worms and other invertebrates pervasively use dsRNA for signalling. Additionally, worms, unlike vertebrates and insects, encode RNA-dependent RNA polymerases that generate dsRNA from ssRNA substrates, enabling amplification of small RNA production. Especially in dsRNA biology, specifying the model organism is essential to avoid confusion about potential effects in humans.

• Similarly, the authors should specify "in C. elegans" in the sentence "Therefore, we propose that the import of extracellular dsRNA into the germline tunes intracellular pathways that cause heritable RNA silencing." This is important because C. elegans small RNA pathways differ significantly from those in other organisms, particularly in the PIWI-interacting RNA (piRNA) pathways, which depend on dsRNA in C. elegans but uses ssRNA in vertebrates. Specification is crucial to prevent misinterpretation by the reader. It is well understood that mechanisms of transgenerational inheritance that operate in nematodes or plants are not conserved in mammals.

• The first sentence of the discussion, "Our analyses suggest a model for ...", would also benefit from specifying "in C. elegans". The same applies to the figure captions. Clarification of the model organism should be added to the first sentence, especially in Figure 1.

Reviewer #2 (Public review):

Summary:

RNAs can function across cell borders and animal generations as sources of epigenetic information for development and immunity. The specific mechanistic pathways how RNA travels between cells and progeny remains an open question. Here, Shugarts, et al. use molecular genetics, imaging, and genomics methods to dissect specific RNA transport and regulatory pathways in the C. elegans model system. Larvae ingesting double stranded RNA is noted to not cause continuous gene silencing throughout adulthood. Damage of neuronal cells expressing double stranded target RNA is observed to repress target gene expression in the germline. Exogenous supply of short or long double stranded RNA required different genes for entry into progeny. It was observed that the SID-1 double-stranded RNA transporter showed different expression over animal development. Removal of the sid-1 gene caused upregulation of two genes, the newly described sid-1-dependent gene sdg-1 and sdg-2. Both genes were observed to also be negatively regulated by other small RNA regulatory pathways. Strikingly, loss then gain of sid-1 through breeding still caused variability of sdg-1 expression for many, many generations. SDG-2 protein co-localizes with a Z-granule marker, an intracellular site for heritable RNA silencing machinery. Collectively, sdg-1 presents a model to study how extracellular RNAs can buffer gene expression in germ cells and other tissues.

Strengths:

(1) Very clever molecular genetic methods and genomic analyses, paired with thorough genetics, were employed to discover insights into RNA transport, sdg-1 and sdg-2 as sid-1-dependent genes, and sdg-1's molecular phenotype.

(2) The manuscript is well cited, and figures reasonably designed.

(3) The discovery of the sdg genes being responsive to the extracellular RNA cell import machinery provides a model to study how exogenous somatic RNA is used to regulate gene expression in progeny. The discovery of genes within retrotransposons stimulates tantalizing models how regulatory loops may actually permit the genetic survival of harmful elements.

Weaknesses:

(1) As presented, the manuscript is incredibly broad, making it challenging to read and consider the data presented. This concern is exemplified in the model figure, that requires two diagrams to summarize the claims made by the manuscript.

(2) The large scope of the manuscript denies space to further probe some of the ideas proposed. The first part of the manuscript, particularly Figures 1 and 2, presents data that can be caused by multiple mechanisms, some of which the authors describe in the results but do not test further. Thus, portions of the results text come across as claims that are not supported by the data presented.

(3) The manuscript focuses on the genetics of SDGs but not the proteins themselves. Few descriptions of the SDGs functions are provided nor is it clarified why only SDG-1 was pursued in imaging and genetic experiments. Additionally, the SDG-1 imaging experiments could use additional localization controls.

Author response:

The following is the authors’ response to the current reviews.

Reviewer #1 (Public review):

Summary:

In the manuscript "Intergenerational transport of double-stranded RNA limits heritable epigenetic changes," Shugarts and colleagues investigate intergenerational dsRNA transport in the nematode C. elegans. By inducing oxidative damage, they block dsRNA import into cells, which affects heritable gene regulation in the adult germline (Fig. 2). They identify a novel gene, sid-1-dependent gene-1 (sdg-1), upregulated upon SID-1 inhibition (Fig. 3). Both transient and genetic depletion of SID-1 lead to the upregulation of sdg-1 and a second gene, sdg-2 (Fig. 5). Interestingly, while sdg-1 expression suggests a potential role in dsRNA transport, neither its overexpression nor loss-of-function impacts dsRNA-mediated silencing in the germline (Fig. 7).

Strengths:

• The authors employ a robust neuronal stress model to systematically explore SID-1 dependent intergenerational dsRNA transport in C. elegans.

• They discover two novel SID-1-dependent genes, sdg-1 and sdg-2.

• The manuscript is well-written and addresses the compelling topic of dsRNA signaling in C. elegans.

Weaknesses:

• The molecular mechanism downstream of SDG-1 remains unclear. Testing whether sdg-2 functions redundantly with sdg-1could provide further insights.

• SDG-1 dependent genes in other nematodes remain unknown.

We thank the reviewer for highlighting the strengths of the work along with a couple of the interesting future directions inspired by the reported discoveries. The restricted presence of genes encoding SDG-1 and its paralogs within retrotransposons suggests intriguing evolutionary roles for these proteins. Future work could examine whether such fast-evolving or newly evolved proteins with potential roles in RNA regulation are more broadly associated with retrotransposons. Multiple SID-1-dependent proteins (including SDG-1 and SDG-2) could act together to mediate downstream effects. This possibility can be tested using combinatorial knockouts and overexpression strains. Both future directions have the potential to illuminate the evolutionarily selected roles of dsRNA-mediated signaling through SID-1, which remain a mystery.

Reviewer #2 (Public review):

Summary:

RNAs can function across cell borders and animal generations as sources of epigenetic information for development and immunity. The specific mechanistic pathways how RNA travels between cells and progeny remains an open question. Here, Shugarts, et al. use molecular genetics, imaging, and genomics methods to dissect specific RNA transport and regulatory pathways in the C. elegans model system. Larvae ingesting double-stranded RNA is noted to not cause continuous gene silencing throughout adulthood. Damage of neuronal cells expressing double-stranded target RNA is observed to repress target gene expression in the germline. Exogenous short or long double-stranded RNA required different genes for entry into progeny. It was observed that the SID-1 double-stranded RNA transporter showed different expression over animal development. Removal of the sid-1 gene caused upregulation of two genes, the newly described sid-1-dependent gene sdg-1 and sdg-2. Both genes were observed to be negatively regulated by other small RNA regulatory pathways. Strikingly, loss then gain of sid-1 through breeding still caused variability of sdg-1 expression for many, many generations. SDG-2 protein co-localizes with germ granules, intracellular sites for heritable RNA silencing machinery. Collectively, sdg-1 presents a model to study how extracellular RNAs can buffer gene expression in germ cells and other tissues.

Strengths:

(1) Very cleaver molecular genetic methods and genomic analyses, paired with thorough genetics, were employed to discover insights into RNA transport, sdg-1 and sdg-2 as sid-1-dependent genes, and sdg-1's molecular phenotype.

(2) The manuscript is well cited, and figures reasonably designed.

(3) The discovery of the sdg genes being responsive to the extracellular RNA cell import machinery provides a model to study how exogenous somatic RNA is used to regulate gene expression in progeny. The discovery of genes within retrotransposons stimulates tantalizing models how regulatory loops may actually permit the genetic survival of harmful elements.

Weaknesses:

(1) The manuscript is broad, making it challenging to read and consider the data presented. Of note, since the original submission, the authors have improved the clarity of the writing and presentation.

Comments on revised version:

This reviewer thanks the authors for their efforts in revising the manuscript. In their rebuttal, the authors acknowledged the broad scope of their manuscript. I concur. While I still think the manuscript is a challenge to read due to its expansive nature, the current draft is substantially improved when compared to the previous one. This work will contribute to our general knowledge of RNA biology, small RNA regulatory pathways, and RNA inheritance.

We thank the reviewer for highlighting the strengths of the manuscript and for helping us improve the presentation of our results and discussion.

The following is the authors’ response to the original reviews.

Public Reviews:

Reviewer #1 (Public review):

Summary:

In the manuscript "Intergenerational transport of double-stranded RNA limits heritable epigenetic changes" Shugarts and colleagues investigate intergenerational dsRNA transport in the nematode C. elegans. They induce oxidative damage in worms, blocking dsRNA import into cells (and potentially affecting the worms in other ways). Oxidative stress inhibits dsRNA import and the associated heritable regulation of gene expression in the adult germline (Fig. 2). The authors identify a novel gene, sid-1-dependent gene-1 (sdg-1), which is induced upon inhibition of SID-1 (Fig. 3). Both transient inhibition and genetic depletion of SID-1 lead to the upregulation of sdg-1 and a second gene, sdg-2 (Fig. 5). The expression of SDG-1 is variable, potentially indicating buffering regulation. While the expression of Sdg-1 could be consistent with a role in intergenerational transport of dsRNA, neither its overexpression nor loss-of-function impacts dsRNA-mediated silencing (Fig. 7) in the germline. It would be interesting to test if sdg-2 functions redundantly.

In summary, the authors have identified a novel worm-specific protein (sdg-1) that is induced upon loss of dsRNA import via SID-1, but is not required to mediate SID-1 RNA regulatory effects.

We thank the reviewer for highlighting our findings on SDG-1. We found that oxidative damage in neurons enhanced dsRNA transport into the germline and/or subsequent silencing.

Remaining Questions:

• The authors use an experimental system that induces oxidative damage specifically in neurons to release dsRNAs into the circulation. Would the same effect be observed if oxidative damage were induced in other cell types?

It is possible that oxidative damage of other tissues using miniSOG (as demonstrated in Xu and Chisholm, 2016) could also enhance the release of dsRNA into the circulation from those tissues. However, future experiments would be needed to test this empirically because it is also possible that the release of dsRNA depends on physiological properties (e.g., the molecular machinery promoting specific secretion) that are particularly active in neurons. We chose to use neurons as the source of dsRNA because by expressing dsRNA in a variety of tissues, neurons appeared to be the most efficient at the export of dsRNA as measured using SID-1-dependent silencing in other tissues (Jose et al., PNAS, 2009).

• Besides dsRNA, which other RNAs and cellular products (macromolecules and small signalling molecules) are released into the circulation that could affect the observed changes in germ cells?

We do not yet know all the factors that could be released either in naive animals or upon oxidative damage of neurons that influence the uptake of dsRNA into other tissues. The dependence on SID-1 for the observed enhancement of silencing (Fig. 2) shows that dsRNA is necessary for silencing within the germline. Whether this import of dsRNA occurs in conjunction with other factors (e.g., the uptake of short dsRNA along with yolk into oocytes (Marré et al., PNAS, 2016)) before silencing within the germline will require further study. A possible approach could be the isolation of extracellular fluid (Banse and Hunter, J Vis Exp., 2012) followed by characterization of its contents. However, the limited material available using this approach and the difficulty in avoiding contamination from cellular damage by the needle used for isolating the material make it challenging.

• SID-1 modifies RNA regulation within the germline (Fig. 7) and upregulates sdg-1 and sdg-2 (Fig. 5). However, SID-1's effects do not appear to be mediated via sdg-1. Testing the role of sdg-2 would be intriguing.

We observe the accumulation of sdg-1 and sdg-2 RNA in two different mutants lacking SID-1, which led us to conservatively focus on the analysis of one of these proteins for this initial paper. We expect that more sensitive analyses of the RNA-seq data will likely reveal additional genes regulated by SID-1. With the ability to perform multiplexed genome-editing, we hope in future work to generate strains that have mutations in many SID-1-dependent genes to recapitulate the defects observed in sid-1(-) animals. Indeed, as surmised by the reviewer, we are focusing on sdg-2 as the first such SID-1-dependent gene to analyze using mutant combinations.

• Are sdg-1 or sdg-2 conserved in other nematodes or potentially in other species? appears to be encoded or captured by a retro-element in the C. elegans genome and exhibits stochastic expression in different isolates. Is this a recent adaptation in the C. elegans genome, or is it present in other nematodes? Does loss-of-function of sdg-1 or sdg-2 have any observable effect?

Clear homologs of SDG-1 and SDG-2 are not detectable outside of C. elegans. Consistent with the location of the sdg-1 gene within a Cer9 retrotransposon that appears to have integrated only within the C. elegans genome, sequence conservation between the genomes of related species is only observed outside the region of the retrotransposon (see Author response image 1, screenshot from UCSC browser). There were no obvious defects detected in animals lacking sdg-1 (Fig. 7) or in animals lacking sdg-2 (data not shown). It is possible that further exploration of both mutants and mutant combinations lacking additional SID-1-dependent genes would reveal defects. We also plan to examine these mutants in sensitized genetic backgrounds where one or more members of the RNA silencing pathway have been compromised.

Clarification for Readability:

To enhance readability and avoid misunderstandings, it is crucial to specify the model organism and its specific dsRNA pathways that are not conserved in vertebrates:

We agree with the reviewer and thank the reviewer for the specific suggestions provided below. To take the spirit of the suggestion to heart we have instead changed the title of our paper to clearly signal that the entire study only uses C. elegans. We have titled the study ‘Intergenerational transport of double-stranded RNA in C. elegans can limit heritable epigenetic changes’

• In the first sentence of the paragraph "Here, we dissect the intergenerational transport of extracellular dsRNA ...", the authors should specify "in the nematode C. elegans". Unlike vertebrates, which recognise dsRNA as a foreign threat, worms and other invertebrates pervasively use dsRNA for signalling. Additionally, worms, unlike vertebrates and insects, encode RNA-dependent RNA polymerases that generate dsRNA from ssRNA substrates, enabling amplification of small RNA production. Especially in dsRNA biology, specifying the model organism is essential to avoid confusion about potential effects in humans.

We agree with most statements made by the reviewer, although whether dsRNA is exclusively recognized as a foreign threat by all vertebrates of all stages remains controversial. Our changed title now eliminates all ambiguity regarding the organism used in the study.

• Similarly, the authors should specify "in C. elegans" in the sentence "Therefore, we propose that the import of extracellular dsRNA into the germline tunes intracellular pathways that cause heritable RNA silencing." This is important because C. elegans small RNA pathways differ significantly from those in other organisms, particularly in the PIWI-interacting RNA (piRNA) pathways, which depend on dsRNA in C. elegans but uses ssRNA in vertebrates. Specification is crucial to prevent misinterpretation by the reader. It is well understood that mechanisms of transgenerational inheritance that operate in nematodes or plants are not conserved in mammals.

The piRNAs of C. elegans are single-stranded but are encoded by numerous independent genes throughout the genome. The molecules used for transgenerational inheritance of epigenetic changes that have been identified thus far are indeed different in different organisms. However, the regulatory principles required for transgenerational inheritance are general (Jose, eLife, 2024). Nevertheless, we have modified the title to clearly state that the entire study is using C. elegans.

• The first sentence of the discussion, "Our analyses suggest a model for ...", would also benefit from specifying "in C. elegans". The same applies to the figure captions. Clarification of the model organism should be added to the first sentence, especially in Figure 1.

With the clarification of the organism used in the title, we expect that all readers will be able to unambiguously interpret our results and the contexts where they apply.

Reviewer #2 (Public review):

Summary:

RNAs can function across cell borders and animal generations as sources of epigenetic information for development and immunity. The specific mechanistic pathways how RNA travels between cells and progeny remains an open question. Here, Shugarts, et al. use molecular genetics, imaging, and genomics methods to dissect specific RNA transport and regulatory pathways in the C. elegans model system. Larvae ingesting double stranded RNA is noted to not cause continuous gene silencing throughout adulthood. Damage of neuronal cells expressing double stranded target RNA is observed to repress target gene expression in the germline. Exogenous supply of short or long double stranded RNA required different genes for entry into progeny. It was observed that the SID-1 double-stranded RNA transporter showed different expression over animal development. Removal of the sid-1 gene caused upregulation of two genes, the newly described sid-1-dependent gene sdg-1 and sdg-2. Both genes were observed to also be negatively regulated by other small RNA regulatory pathways. Strikingly, loss then gain of sid-1 through breeding still caused variability of sdg-1 expression for many, many generations. SDG-2 protein co-localizes with a Z-granule marker, an intracellular site for heritable RNA silencing machinery. Collectively, sdg-1 presents a model to study how extracellular RNAs can buffer gene expression in germ cells and other tissues.

We thank the reviewer for highlighting our findings and underscoring the striking nature of the discovery that mutating sid-1 using genome-editing resulted in a transgenerational change that could not be reversed by changing the sid-1 sequence back to wild-type.

Strengths:

(1) Very clever molecular genetic methods and genomic analyses, paired with thorough genetics, were employed to discover insights into RNA transport, sdg-1 and sdg-2 as sid-1-dependent genes, and sdg-1's molecular phenotype.

(2) The manuscript is well cited, and figures reasonably designed.

(3) The discovery of the sdg genes being responsive to the extracellular RNA cell import machinery provides a model to study how exogenous somatic RNA is used to regulate gene expression in progeny. The discovery of genes within retrotransposons stimulates tantalizing models how regulatory loops may actually permit the genetic survival of harmful elements.

We thank the reviewer for the positive comments.

Weaknesses:

(1) As presented, the manuscript is incredibly broad, making it challenging to read and consider the data presented. This concern is exemplified in the model figure, that requires two diagrams to summarize the claims made by the manuscript.

RNA interference (RNAi) by dsRNA is an organismal response where the delivery of dsRNA into the cytosol of some cell precedes the processing and ultimate silencing of the target gene within that cell. These two major steps are often not separately considered when explaining observations. Yet, the interpretation of every RNAi experiment is affected by both steps. To make the details that we have revealed in this work for both steps clearer, we presented the two models separated by scale - organismal vs. intracellular. We agree that this integrative manuscript appears very broad when the many different findings are each considered separately. The overall model revealed here forms the necessary foundation for the deep analysis of individual aspects in the future.

(2) The large scope of the manuscript denies space to further probe some of the ideas proposed. The first part of the manuscript, particularly Figures 1 and 2, presents data that can be caused by multiple mechanisms, some of which the authors describe in the results but do not test further. Thus, portions of the results text come across as claims that are not supported by the data presented.

We agree that one of the consequences of addressing the joint roles of transport and subsequent silencing during RNAi is that the scope of the manuscript appears large. We had suggested multiple interpretations for specific observations in keeping with the need for further work. To avoid any misunderstandings that our listing of possible interpretations be taken as claims by the reader, we have followed the instructions of the reviewer (see below) and moved some of the potential explanations we raised to the discussion section.

(3) The manuscript focuses on the genetics of SDGs but not the proteins themselves. Few descriptions of the SDGs functions are provided nor is it clarified why only SDG-1 was pursued in imaging and genetic experiments. Additionally, the SDG-1 imaging experiments could use additional localization controls.

We agree that more work on the SDG proteins will likely be informative, but are beyond the scope of this already expansive paper. We began with the analysis of SDG-1 because it had the most support as a regulator of RNA silencing (Fig. 5f). Indeed, in other work (Lalit and Jose, bioRxiv, 2024), we find that AlphaFold 2 predicts the SDG-1 protein to be a regulator of RNA silencing that directly interacts with the dsRNA-editing enzyme ADR-2 and the endonuclease RDE-8. Furthermore, we expect that more sensitive analyses of the RNA-seq data are likely to reveal additional genes regulated by SID-1. Using multiplexed genome editing, we hope to generate mutant combinations lacking multiple sdg genes to reveal their function(s).

We agree that given the recent discovery of many components of germ granules, our imaging data does not have sufficient resolution to discriminate between them. We have modified our statements and our model regarding the colocalization of SDG-1 with Z-granules to indicate that the overlapping enrichment of SDG-1 and ZNFX-1 in the perinuclear region is consistent with interactions with other nearby granule components.

Recommendations for the authors:

Reviewer #2 (Recommendations for the authors):

Major

(1) As presented, the manuscript is almost two manuscripts combined into one. This point is highlighted in Figure 7h, which basically presents two separate models. The key questions addressed in the manuscript starts at Figure 3. Figures 1 and 2 are interesting observations but require more experiments to define further. For example, as the Results text describes for Figure 1, "These differences in the entry of ingested dsRNA into cells and/or subsequent silencing could be driven by a variety of changes during development. These include changes in the uptake of dsRNA into the intestine, distribution of dsRNA to other tissues from the intestine, import of dsRNA into the germline, and availability of RNA silencing factors within the germline." Presenting these (reasonable) mechanistic ideas detracted from the heritable RNA epigenetic mechanism explored in the later portion of the manuscript. There are many ways to address this issue, one being moving Figures 1 and 2 to the Supplement to focus on SID-1 related pathways.

Since this manuscript addresses the interaction between intercellular transport of dsRNA and heritable epigenetic changes, it was necessary to establish the possible route(s) that dsRNA could take to the germline before any inference could be made regarding heritable epigenetic changes. As suggested below (pt. 2), we have now moved the alternatives we enumerated as possible explanations for some experimental results (e.g., for the differences quoted here) to the discussion section.

(2) The manuscript includes detailed potential interpretations in the Results, making them seem like claims. Here is an example:

"Thus, one possibility suggested by these observations is that reduction of sdg-1 RNA via SID-1 alters the amount of SDG-1 protein, which could interact with components of germ granules to mediate RNA regulation within the germline of wild-type animals."

This mechanism is a possibility, but placing these ideas in the citable results makes it seem like an overinterpretation of imaging data. This text and others should be in the Discussion, where speculation is encouraged. Results sections like this example and others should be moved to the discussion.

We have rephrased motivating connections between experiments like the one quoted above and also moved such text to the discussion section wherever possible.

(3) A paragraph describing the SDG proteins will be helpful. Homologs? Conserved protein domains? mRNA and/or protein expression pattern across worm, not just the germline? Conservation across Caenorhabditis sp? These descriptions may help establish context why SDG-1 localizes to Z-granules.

We have now added information about the conservation of the sdg-1 gene in the manuscript. AlphaFold predicts domains with low confidence for the SDG-1 protein, consistent with the lack of conservation of this protein (AlphaFold requires multiple sequence alignments to predict confidently). In the adult animal, the SDG-1 protein was only detectable in the germline. Future work focused on SDG-1, SDG-2 and other SDG proteins will further examine possible expression in other tissues and functional domains if any. Unfortunately, in multiple attempts of single-molecule FISH experiments using probes against the sdg-1 open reading frame, we were unable to detect a specific signal above background (data not shown). Additional experiments are needed for the sensitive detection of sdg-1 expression outside the germline, if any.

(4) Based on the images shown, SDG-1 could be in other nearby granules, such as P granules or mutator foci. Additional imaging controls to rule out these granules/condensates will greatly strengthen the argument that SDG-1 protein localizes to Z-granules specifically.

We have modified the final model to indicate that the perinuclear colocalization is with germ granules broadly and we agree that we do not have the resolution to claim that the observed overlap of SDG-1::mCherry with GFP::ZNFX-1 that we detect using Airyscan microscopy is specifically with Z granules. Our initial emphasis of Z-granule was based on the prior report of SDG-1 being co-immunoprecipitated with the Z-granule surface protein PID-2/ZSP-1. However, through other work predicting possible direct interactions using AlphaFold (Lalit and Jose, bioRxiv, 2024), we were unable to detect any direct interactions between PID-2 and SDG-1. Indeed, many additional granules have been recently reported (Chen et al., Nat. Commun., 2024; Huang et al., bioRxiv 2024), making it possible that SDG-1 has specific interactions with a component of one of the other granules (P, Z, M, S, E, or D) or adjacent P bodies.

Minor

(1) "This entry into the cytosol is distinct from and can follow the uptake of dsRNA into cells, which can rely on other receptors." Awkard sentence. Please revise.

We have now revised this sentence to read “This entry into the cytosol is distinct from the uptake of dsRNA into cells, which can rely on other receptors”

(2) Presumably, the dsRNA percent of the in vitro transcribed RNA is different than the 50 bp oligos that can be reliably annealed by heating and cooling. Other RNA secondary structure possibilities warrant further discussion.

We agree that in vitro transcribed RNA could include a variety of undefined secondary structures in addition to dsRNAs of mixed length. Such structures could recruit or titrate away RNA-binding proteins in addition to the dsRNA structures engaging the canonical RNAi pathway, resulting in mixed mechanisms of silencing. Future work identifying such structures and exploring their impact on the efficacy of RNAi could be informative. We have now added these considerations to the discussion and thank the reviewer for highlighting these possibilities.

Version 2

Abstract

Abstract

RNAs in circulation carry sequence-specific regulatory information between cells in plant, animal, and host-pathogen systems. Such RNA can cross generational boundaries, as evidenced by somatic double-stranded RNA (dsRNA) in the nematode C. elegans silencing genes of matching sequence in progeny. Here we dissect the intergenerational path taken by dsRNA from parental circulation and discover that cytosolic import through the dsRNA importer SID-1 in the parental germline and/or developing progeny varies with developmental time and dsRNA substrates. Loss of SID-1 enhances initiation of heritable RNA silencing within the germline and causes changes in the expression of the sid-1-dependent gene sdg-1 that last for more than 100 generations after restoration of SID-1. The SDG-1 protein is enriched in perinuclear germ granules required for heritable RNA silencing but is expressed from a retrotransposon targeted by such silencing. This auto-inhibitory loop suggests how retrotransposons could persist by hosting genes that regulate their own silencing.

Main text

RNAs released into circulation can act as intercellular messages that are used for gene regulation in distant cells. Examples include secretion of small RNAs within exosomes in response to pathogenic fungal infection in Arabidopsis (Cai et al., 2018), virus-like proteins with their coding mRNAs in developing Drosophila (Ashley et al., 2018) and mice (Pastuzyn et al., 2018), microRNAs from adipose tissue in mice (Thomou et al., 2017) and small RNAs from the epididymis in mice (Sharma et al., 2016; Chen et al., 2016; Conine et al., 2018; Sharma et al., 2018). Such extracellular RNAs have also been detected in humans, but their roles in gene regulation remain unclear despite their use as diagnostic markers for diseases (reviewed in Das et al., 2019). Furthermore, the recent development of double-stranded RNA (dsRNA)-based drugs (reviewed in Setten et al., 2019; Hu et al., 2020; Mullard, 2022) that can silence genes of matching sequence through RNA interference (Fire et al., 1998) has heightened interest in understanding the import of dsRNA into cells. A conserved dsRNA-selective importer, SID-1 (Winston et al., 2002; Feinberg and Hunter, 2003; Shih et al., 2009), is required for the import of extracellular dsRNA into the cytosol of any cell in the nematode C. elegans. This entry into the cytosol is distinct from the uptake of dsRNA into cells, which can rely on other receptors (e.g., SID-2 for uptake from the intestinal lumen (Winston et al., 2007; McEwan et al., 2012). SID-1 has two homologs in mammals – SIDT1 and SIDT2. Although similar cytosolic entry of dsRNA using these mammalian homologs of SID-1 is supported by studies in mice reporting entry of viral dsRNA through SIDT2 (Nguyen et al., 2017), enhanced dsRNA uptake when SIDT1 is overexpressed in vitro (Nguyen et al., 2019), and uptake of ingested dsRNA into cells through SIDT1 (Chen et al., 2021), alternative roles for SIDT1 and/or SIDT2 in the uptake of cholesterol have also been proposed (Méndez-Acevedo et al., 2017).

Secretion of dsRNA from C. elegans tissues expressing dsRNA from transgenes has been inferred based upon the SID-1-dependent silencing of matching genes in other tissues (Winston et al., 2002; Jose et al., 2009). Secreted dsRNA from neurons can silence genes of matching sequence in most somatic cells (Ravikumar et al., 2019) and within the germline (Devanapally et al., 2015). Extracellular dsRNA delivered into parental circulation by injection or ingestion also enters the germline and can cause silencing of matching genes in progeny (Fire et al., 1998; Timmons and Fire, 1998; Grishok et al., 2000; Marré et al., 2016; Wang and Hunter, 2017). In every case, the entry of dsRNA into the cytosol dictates when and where the processing of extracellular dsRNA can begin. Such intergenerational transport of RNA is an attractive mechanism for explaining endogenous, gene-specific effects in progeny that could occur in response to changes in somatic tissues of parents. However, which conditions induce transport of dsRNA into the germline, when during development this transport occurs, and what regulatory consequences ensue in progeny upon uptake of extracellular dsRNA from parents are all unknown. Despite this lack of knowledge, the analysis of transgenerational gene silencing triggered by dsRNA has revealed that a class of small RNAs called 22G RNA made using the mRNA targeted by dsRNA and bound by Argonaute proteins in the germline (Buckley et al., 2012; Xu et al., 2018) is necessary for observing silencing in every generation.

Timed dsRNA injection into animals or timed dsRNA ingestion by animals has been the only way to study mechanisms of dsRNA transport throughout C elegans development since no tools to induce dsRNA secretion from cells have been developed. Injection of dsRNA into adult sid-1(-) animals demonstrated that extracellular dsRNA can be directly transmitted to progeny without entry into the cytosol (Marré et al., 2016; Wang and Hunter, 2017). This intergenerational transmission of dsRNA in sid-1(-) adult animals requires the yolk receptor RME-2 (Grant and Hirsh, 1999) and is independent of parental 22G RNA production because dsRNA cannot enter the cytosol in the parent. In further support of the initial intergenerational transport of extracellular dsRNA or dsRNA-derived signals not requiring 22G RNAs, gfp-dsRNA from parents that lack gfp sequences are transported to progeny (Figure 1E in Marré et al., 2016). Thus, the silencing signals transported from parent to progeny include the extracellular dsRNA and its derived silencing signals, independent of 22G RNAs. However, 22G RNAs are required for silencing and they may be used for intergenerational transport in subsequent generations.

Gene silencing by ingested dsRNA during larval development can fail to persist into adulthood. Silencing in the germline was measured after exposure of gtbp-1::gfp animals to bacteria expressing dsRNA by imaging separate cohorts at indicated stages (A) or day 3 of adulthood (B). (A and B, left) Schematics depicting stages and durations of exposure to dsRNA. (A and B, right) GFP intensity (a.u.) in gtbp-1::gfp animals quantified in germ cells (larvae) or eggs in utero (adults) (A) or in day 3 adult (a3) animals (B) after exposure to control dsRNA (black) or gfp-dsRNA (red). The numbers of animals scored at each stage (n) are indicated. Asterisks indicate P < 0.05 with Bonferroni correction using Mann-Whitney U test for two-sided comparisons between animals exposed to control or gfp-dsRNA. Also see Figure 1–figure supplement 1.

Uptake of dsRNA into the proximal germline by RME-2 is required for silencing during early adulthood.

(A, left) Schematic depicting continuous exposure of gtbp-1::gfp P0 animals, starting at the L4 stage, and their F1 progeny to bacteria expressing dsRNA, followed by imaging of animals at the indicated stages. (right) Quantification of GTBP-1::GFP intensity (arbitrary units, a.u.) in representative germ cells (larvae) or embryos in utero (adults) of F1 animals at indicated stages after P0 and F1 exposure to control (dark grey) or gfp-dsRNA (red). Numbers of animals scored at each stage (n) are indicated. Asterisks indicate P < 0.05 with Bonferroni correction using Mann-Whitney U test for two-sided comparisons between animals exposed to L4440 or gfp-dsRNA. (B) Silencing of gtbp-1::gfp in germlines injected with duplex buffer (buffer) or in vitro transcribed gfp-dsRNA in duplex buffer during the first day of adulthood and scored for silencing 24, 36 and 48 h post injection. The numbers of animals out of 5 injected with each injection mix that exhibited silencing of both gonad arms are indicated for each time point. Animals injected with buffer never exhibited silencing in either gonad arm. (C) Hermaphrodite animals of the L4 stage (left) or young adult stage (24 hour post L4, right) of the indicated genotypes were fed unc-22-dsRNA expressed in bacteria for 24 hours (red). Hermaphrodite self-progeny of fed animals were scored for unc-22 silencing (strong, black; weak, grey). Numbers of fed P0 parents and scored F1 progeny (P0; F1 n) are as indicated. Asterisk indicates P < 0.05 with Bonferroni correction using χ2 test. Previously generated rme-2(-) animals were used in this assay (DH1390). (D) Expression of RME-2. (top) Schematic showing insertion of wrmScarlet (rme-2(jam119[wrmScarlet])) at the rme-2 locus. Scale bar, 1 kb. (bottom) Brightfield and fluorescence images of a rme-2(jam119[wrmScarlet]) L4-stage and adult animal (n = 1 confocal plane). Scale bars, 20 μm. (E) Representative fluorescence images of GTBP-1::GFP (black) in the germlines (dashed outline) of day 3 gtbp-1::gfp adult animals after P0 and F1 ingestion of control dsRNA (left) or gfp-dsRNA (right) up to the first day of adulthood. Numbers of animals imaged (n) and the percentages of animals exhibiting the depicted expression patterns are shown. Scale bars, 50 μm.

All dsRNAs, regardless of length, have been assumed to be equivalent substrates for entry into the cytosol. This assumption is supported by the uptake of a variety of dsRNA substrates when C. elegans SID-1 is overexpressed in heterologous Drosophila S2 cells (Shih et al., 2009). Yet, two key observations suggest that dsRNA can take multiple routes in vivo before SID-1-dependent entry into the cytosol. One, even in the presence of SID-1, dsRNA ingested during early adulthood requires RME-2 to cause silencing in progeny (Marré et al., 2016). Two, 50-bp fluorescently labeled dsRNA requires RME-2 for entry from parental circulation into oocytes (Marré et al., 2016). Establishing the contexts for the use of different modes of dsRNA transport is crucial for understanding the processes regulated by endogenous extracellular dsRNA.

Here, we dissect the intergenerational transport of extracellular dsRNA and discover a role for this mechanism in modulating RNA regulation within the germline. Extracellular dsRNA is transported with developmental and substrate specificity from parental circulation to progeny, and its release from neurons can be enhanced using light-induced oxidative damage. Blocking dsRNA import into the cytosol of all cells revealed heritable changes in gene expression and led to the identification of sid-1-dependent gene-1 (sdg-1). The sdg-1 coding sequence is located within a retrotransposon that is targeted by RNA silencing in the germline. Yet, the SDG-1 protein colocalizes with regulators of RNA silencing in perinuclear granules within the germline and dynamically enters the nucleus in proximal oocytes and in cells of developing embryos.

Measurements of sdg-1 expression using native mRNA, a translational reporter, or a transcriptional reporter reveal that expression is easily perturbed in different mutants that impact dsRNA-mediated gene regulation. Expression varies between the two gonad arms of wild-type animals, and different mutant isolates can show an increase or decrease in expression, indicative of a loss of buffered gene expression within the germline. However, consistent with an overall role for SDG-1 (and potentially other SDGs) in promoting RNA silencing, either loss of SID-1 or overexpression of SDG-1 enhances piRNA-mediated silencing within the germline initiated by mating. Therefore, we propose that the import of extracellular dsRNA into the germline tunes intracellular pathways that cause heritable RNA silencing.

Requirements for the entry of extracellular dsRNA into the germline change during development

A convenient method for the delivery of extracellular dsRNA into C. elegans at various times during larval development is the expression of dsRNA in the bacteria that the animals ingest as food (Timmons and Fire, 1998). To determine when ingested dsRNA can enter the germline and cause silencing, we exposed developing animals with a ubiquitously expressed protein (GTBP-1) tagged with GFP to bacteria that express gfp-dsRNA. Silencing was detectable within the germline from the second larval stage (L2) onwards (Figure 1A and Figure 1–figure supplement 1A), but either exposure to ingested dsRNA beyond the fourth larval stage (L4) (Figure 1B) or injection of dsRNA into the 1-day old adult germline (Figure 1–figure supplement 1B) was required to observe silencing in the germline of 3-day old adults. Combined with the need for exposure to dsRNA after the L4 stage (Marré et al., 2016; Wang and Hunter, 2017) for silencing in progeny, even for just 24 hours (Figure 1–figure supplement 1C), these observations suggest that detectable silencing within the germline is possible during early development without initiating heritable RNA silencing. One possible explanation for this observation could be that both RNAs derived from the imported dsRNA and downstream silencing signals are continually diluted by the proliferation of germ cells. Heritable silencing by dsRNA ingested from the L4 stage to the first day of adulthood likely relies on entry of dsRNA into the proximal germline because silencing of a somatic gene in progeny after parental ingestion of dsRNA during this period required RME-2 (Figure 1–figure supplement 1C), which is enriched in the proximal germline (Figure 1–figure supplement 1D; Grant and Hirsh, 1999), and some gtbp-1::gfp animals exposed to gfp-dsRNA up to the first day of adulthood showed more silencing in the proximal germline (Figure 1–figure supplement 1E).

Thus, these results reveal three periods of germline development that can be broadly distinguished based on the response to ingested dsRNA: (1) from the first larval to the fourth larval stage when exposure to dsRNA does not result in maximal silencing within the germline in adults (Figure 1B); (2) from the fourth larval stage to early adulthood when entry of dsRNA primarily occurs in the proximal germline through RME-2 (Figure 1–figure supplement 1C and 1E); and (3) later adulthood when germline silencing by ingested dsRNA is maximal (Figure 1B) and ingested dsRNA can effectively silence progeny independent of RME-2 (Figure 1–figure supplement 1C; Wang and Hunter, 2017).

Oxidative damage in neurons expressing dsRNA enhances silencing in the germline by neuronal dsRNA

Another approach for delivering extracellular dsRNA into the germline that better mimics dsRNA transport between cells is the secretion of dsRNA from neurons (Devanapally et al., 2015). However, the extent of such secretion throughout development is unpredictable. To modulate the secretion of dsRNA from somatic cells into parental circulation during development, we adapted an optogenetic approach for damaging somatic cells (Xu and Chisholm, 2016). Specifically, we generated animals that express the mini singlet oxygen generator (miniSOG) protein in neurons and exposed them to blue light. While animals expressing miniSOG from a single-copy transgene did not show an appreciable defect when compared with wild-type animals, those expressing miniSOG from a multi-copy transgene were paralyzed (Figure 2–figure supplement 2A and 1B, top) and had visibly damaged neurons (Figure 2–figure supplement 2B, bottom). Using this system, we induced oxidative damage in the neurons of animals that expressed dsRNA under the control of a neuronal promoter and evaluated silencing of target genes with matching sequence expressed in other tissues (Figure 2A). By exposing animals to blue light for 60 minutes at different times during development (Figure 2–figure supplement 2C), we observed SID-1-dependent enhancement in the silencing of the hypodermal gene bli-1 at the adult stage by neuronal bli-1-dsRNA, with maximal silencing when oxidative damage occurred during mid-to-late larval development (Figure 2–figure supplement 2D, light exposure every 6 hours from 42 to 66 hours post L4-stage of parent; Figure 2–figure supplement 2E, ∼2-fold increase from 14.9% to 29.1% in a background with enhanced RNA interference (eri-1(-)) and ∼6-fold increase from ∼1.6% to ∼9.8% in a wild-type background). A similar period of maximal SID-1-dependent enhancement of silencing was also observed when neurons expressing gfp-dsRNA were damaged and silencing of a two-gene operon that expresses two fluorescent proteins, mCherry::H2B and GFP::H2B, in the germline was measured (Figure 2B-2D, and Figure 2–figure supplement 2F – 48 to 60 hours post L4-stage of parent; sid-1(-) allele (jam80[nonsense]) is depicted in Figure 2–figure supplement 2). While silencing of gfp::h2b was observed throughout the germline, silencing of the other cistron mCherry::h2b was also observed sometimes, albeit restricted to specific regions of the germline. Silencing of mCherry::h2b was most frequent in the proximal germline and was not observed in any other region without silencing in the proximal germline (proximal germline – 57%, distal germline – 47%, sperm – 29%, Figure 2D), likely due to reduction of mCherry::h2b::gfp::h2b pre-mRNA (Guang et al., 2008) in those regions. Consistently, the silencing of both gfp::h2b and mCherry::h2b was eliminated in the absence of the nuclear Argonaute HRDE-1 (Figure 2–figure supplement 2G). The pattern of mCherry::h2b silencing is similar to the spatial pattern observed for the RME-2-dependent entry of dsRNA delivered into parental circulation (Marré et al., 2016) and is consistent with the pattern of target mRNA degradation in the germline by extracellular dsRNA (Ouyang et al., 2022). However, silencing of gfp::h2b in the germline by neuronal dsRNA did not show a detectable dependence on RME-2 (Figure 2–figure supplement 2H - difficulty in obtaining transgenic animals that also lack RME-2 resulted in a low sample size for this experiment).

Oxidative damage of neurons enhances gene silencing by neuronal dsRNA in the adult germline. (A) Schematic illustrating exposure to blue light of animals expressing a singlet oxygen generator (miniSOG) and gfp-dsRNA in neurons, and subsequent release of dsRNA. Such extracellular dsRNA is expected to enter the cytosol of the germline through the dsRNA importer SID-1 and silence gfp::h2b mRNA from a two-gene operon that expresses mCherry::h2b and gfp::h2b as part of a single pre-mRNA. (B-D) Images of single gonad arms in adult animals with the two-gene operon (mex-5p::mCherry::h2b::gfp::h2b) showing fluorescence (black) of mCherry::H2B (magenta outline) or of GFP::H2B (green outline). Punctate autofluorescence from the intestine can also be seen. Numbers of animals assayed (n) and percentages of adult animals with the depicted expression patterns are indicated. Scale bars, 50 μm. (B) mCherry::H2B fluorescence is seen throughout the germline (left) and GFP::H2B fluorescence is seen in the oocytes and in the distal gonad (right). (C) GFP::H2B fluorescence in sid-1(+) and sid-1(-) animals expressing membrane-localized miniSOG (PH::miniSOG) and gfp-dsRNA driven by a neuronal promoter (rgef-1p) from a multi-copy transgene (Ex, jamEx214) without (left) or with (right) exposure to blue light at 48 hours post L4-stage of parent. (D) mCherry::H2B fluorescence in sid-1(+) animals with the transgene Ex. Silencing of mCherry is enhanced in the distal gonad (third row) and sperm (fourth row) after exposing animals to blue light at 48 hours and 54 hours post L4-stage of parent. By region, silencing after exposure to light (right) in the proximal germline (57% = 10 + 18 + 29) > distal germline (47% = 18 + 29) > sperm (29%). Also see Figure 2–figure supplements 1 and 2.

Timed release of neuronal dsRNA by oxidative damage in neurons reveals period of enhanced gene silencing in the soma and germline.

(A) Wild-type animals (left) and animals expressing membrane-tethered mini singlet oxygen generator protein (PH::miniSOG) from an extrachromosomal array (Ex, middle) or a single-copy transgene (Si, right) under a pan-neuronal promoter (rgef-1p) were exposed to blue light for different durations (minutes) and animals were scored for paralysis immediately after exposure (0 h, black) and 24 hours later (24 h, grey). (B) Functional and anatomical evidence for oxidative damage in neurons. (top) Widefield images of animals without (left) and with (right) Ex[rgef-1p::PH::miniSOG] after 5 minutes of blue light exposure. Animals paralyzed in a often appear coiled (right), likely indicative of a defect in neuronal signaling. Scale bar, 100 μm. (bottom) Confocal fluorescence images of neurons in the head region of animals with Ex[rgef-1p::(PH::miniSOG & DsRed)] without (left) and with (right) 30 minutes of blue light exposure showing light-induced changes (black, DsRed fluorescence). Scale bar, 20 μm. (C) Schematic of assay for measuring the impact of oxidative damage in neurons at different times during development on silencing by neuronal dsRNA. For measuring silencing in the hypodermis (top) or germline (bottom), cohorts of animals with Ex[rgef-1p::(PH::miniSOG & bli-1-dsRNA)] (top), or Ex[rgef-1p::(PH::miniSOG & gfp-dsRNA)] obtained by mating males with the array and hermaphrodites with Si[mex-5p::mCherry::h2b::gfp::h2b] (bottom) were exposed to blue light as indicated and scored for bli-1 silencing (top) or imaged (bottom) as stage-matched adults (at ∼96 hours after the fourth larval stage of parent animals). (D) Percentages of eri-1(mg366) (red) or eri-1(mg366); sid-1(qt9) (black) animals silenced when assayed as described in (C, top). Silencing in the absence of exposure to blue light (no light) was also measured for comparison. (E) Percentages of stage-matched animals of the indicated genetic backgrounds with Ex[rgef-1p::(PH::miniSOG & bli-1-dsRNA)] that show bli-1 silencing without (black) or with (blue) a 1-hour exposure to blue light 48 hours after the fourth larval stage of parent animals. The 48 hr time point from (D) is replotted to facilitate comparison. (F) Fractions of animals exhibiting bright (light grey), dim (dark grey) or not detectable (black) mCherry::H2B or GFP::H2B fluorescence in the distal gonad (top), proximal gonad (middle) or sperm (bottom) when assayed as described in (C, bottom). Silencing in the absence of exposure to blue light (no light) was used as the reference. Numbers of animals scored (n), measurements that were not done (nd), significant differences using two-tailed Wilson’s estimates for single proportion compared to wild type (asterisks in (A)) or no light condition (asterisks in (D) and (E)) or χ2 test compared to no light condition (hashes in (F); P < 0.05 with Bonferroni correction), and error bars (95% CI) are indicated. (G and H) Animals homozygous (G) or hemizygous (H) for the mex-5p::mCherry::h2b::gfp::h2b transgene (Figure 2) with or without neuronal gfp-dsRNA (jamEx140) were scored for expression of mCherry and GFP (bright, dim, off) in otherwise wild-type (+), hrde-1(-) (G) or rme-2(-) (H) backgrounds. Animals in (G) also have a dpy-2(e8) mutation linked to the mex-5p::mCherry::h2b::gfp::h2b transgene. Fraction silenced in wild type animals (+) in (G) were calculated with n = 31 for GFP and n = 27 for mCherry. Asterisks indicates P < 0.05 using χ2 test with Bonferroni correction.

Schematics depicting mutations generated in this study

Structures (boxes, exons; lines, introns) and chromosomal locations of genes with mutations generated using Cas9-mediated genome editing. Nonsense mutations (e.g., jam182[nonsense]) with associated amino acid changes (e.g., W161* for tryptophan at position 161 to stop) are indicated with black arrowheads and deletions of coding regions (e.g., jam134[deletion]) are indicated with a dashed line (deleted region) and flanking black arrowheads. Scale bar, 1 kb.

Thus, by modulating the secretion of dsRNA from somatic cells for the first time, we gained two insights into the intercellular transport of dsRNA: (1) oxidative damage of neurons during particular periods in development increases the amount of dsRNA and/or changes the kinds of dsRNA in circulation either because of specific enhancement of secretion or nonspecific spillage; and (2) there is a preference for silencing by neuronal dsRNA in the proximal germline.

Extracellular dsRNA in parental circulation can be transported through multiple routes to cause silencing in progeny

While the characteristics of extracellular dsRNA imported into the germline from ingested bacteria or from neurons are unknown, delivery of chemically defined dsRNA into the extracellular space in C. elegans can be accomplished using microinjection (Fire et al., 1998; Marré et al., 2016). We examined differences, if any, in the entry of in vitro transcribed dsRNA from the extracellular body cavity into the germline during the L4 and adult stages as evidenced by silencing in the progeny of injected animals. Silencing was comparable regardless of whether wild-type or rme-2(-) parents were injected as L4-staged or adult animals (Figure 3A and Figure 3–figure supplement 3A, left; also reported for adults in Wang and Hunter, 2017). However, some dependence on RME-2 for silencing in progeny was discernable when lower concentrations of dsRNA were used (Figure 3–figure supplement 3A, right). This result and previous results demonstrating proximal germline-specific silencing in contexts where silencing is less robust (Figure 1–figure supplement 1E and Figure 2D) are consistent with RME-2-dependent uptake of dsRNA being a route for extracellular dsRNA to enter the germline. The difference in parental RME-2 requirement for silencing in progeny observed for dsRNA ingested (Figure 1–figure supplement 1C) or injected (Figure 3–figure supplement 3A) at the L4 stage could similarly reflect the accumulation of different amounts of dsRNA in parental circulation (e.g., more upon injection than upon ingestion), and/or different kinds of dsRNA (e.g., because of modifications in bacteria or upon transit through the intestine). However, these possibilities could not be easily distinguished because sensitive northern blotting (Choi et al., 2017) revealed that both bacterial and in vitro transcribed dsRNA consist of a complex mix of dsRNAs (Figure 3–figure supplement 3B-1D; consistent with Tabara et al., 2002 and Jain et al., 2020), hereafter called mixed dsRNA.

Transport of dsRNA from parental circulation to progeny occurs through two routes with distinct substrate selectivity.

(A) Hermaphrodite animals of indicated genotypes (in red) were injected in the body cavity with 50-bp unc-22-dsRNA synthesized with a 5’-OH (short dsRNA, left bars) or unc-22-dsRNA with a 5’ triphosphate transcribed from a ∼1.1 kb template (mixed dsRNA, right bars). Hermaphrodite self-progeny of injected animals were scored for unc-22 silencing (fr. Unc-22: strong, black; weak, grey). Numbers of injected parents and scored progeny (P0; F1 n) are indicated. Also see Figure 3–figure supplements 1 and 2. (B) Fluorescence images of progeny from animals with a gfp tag of the ubiquitously expressed gene gtbp-1 (gtbp-1::gfp) that were not injected (left), injected with 50-bp gfp-dsRNA (short dsRNA injection, middle), or injected with dsRNA transcribed from a ∼730-bp template (mixed dsRNA injection, right). Complete silencing is not observed in neurons or in the developing vulva; brackets indicate additional regions with dim GFP fluorescence. Numbers of animals assayed (n) and percentages of L4-staged animals with the depicted expression patterns are indicated. Scale bar, 100 μm. Also see Figure 3–figure supplement 3. (C) Requirements for intergenerational transport of extracellular dsRNA. (top left) Differential Interference Contrast (DIC) and fluorescence images of a developing embryo from an animal injected in the body cavity with 50-bp dsRNA of the same sequence as in (B) and labeled at the 5’ end of the antisense strand with Atto-565. Accumulation within the intestinal lumen (arrowhead), number of embryos imaged (n), and percentage of embryos with depicted pattern of fluorescence are indicated. Scale bar, 20 μm. (top right and bottom) Hermaphrodite animals of the indicated genotypes were injected with short dsRNA (left bars) or mixed dsRNA (right bars) and self-progeny (top right) or cross progeny after mating with wild-type males (bottom) were analyzed as in (A). Cases of no observable silencing are indicated with ‘0’. (D) Schematic summarizing requirements for transport of dsRNA from parental circulation to developing progeny. See text for details. Asterisks in (A) and (C) indicate P < 0.05 with Bonferroni correction using χ2 test.

Requirement of RME-2 for silencing in progeny by dsRNA injected into parents depends on concentration, length, and 5’ modification of dsRNA.

(A) Hermaphrodite animals of indicated genotypes were injected in the body cavity with unc-22-dsRNA (red font) and uninjected F1 progeny were scored for unc-22 silencing (strong, black; weak, grey). Numbers of injected P0 parents and scored F1 progeny (P0; F1 n) are as indicated. (left) L4-staged hermaphrodites were injected with transcribed unc-22-dsRNA at the same concentration as in Figure 3A (1X). (right) Young adult-staged hermaphrodites were injected with transcribed unc-22-dsRNA at ∼0.25X of concentration in Figure 3A. (B and C) Northern blots of bacterial unc-22-dsRNA (unc-22, (B)) or gfp-dsRNA (gfp, (C)) separated alongside empty vector control RNA using fully-denaturing formaldehyde polyacrylamide gel electrophoresis (FDF-PAGE) (Harris et al., 2015). 40-nt digoxigenin (DIG)-labeled oligonucleotides (in blue) were used to probe the 5’ end, middle and 3’ end of the sense (top) and antisense (bottom) strands of the unc-22 (B) and gfp (C) sequences present in the bacterial vectors. A 1-kb DNA ladder was used as a size reference and 5S rRNA was probed as a control for equal loading of total RNA. (D) Northern blot of unc-22-dsRNA transcribed from a ∼1.1-kb template, separated using FDF-PAGE as in (B) and (C), and probed using 40-nt DIG-labeled oligonucleotides complementary to the sense (left) or antisense (right) strands of the unc-22 gene. (E) Polyacrylamide gel stained with ethidium bromide showing 50-nt single-stranded (sense, antisense, 5’P-sense, 5’P-antisense) and 50-bp double-stranded unc-22-RNA (annealed, 5’P-annealed). A 100-bp DNA ladder was run alongside for rough size estimation. 5’-phosphate (5’P) was added using a polynucleotide kinase. (F) Young adult-staged hermaphrodites were injected in the body cavity with short unc-22-dsRNA with 5’-OH (left) or with 5’-phosphate added using a polynucleotide kinase (right) and self-progeny were scored as in (A). Newly generated rme-2(-) animals (AMJ1131) were used in (A) and (F). Comparisons with P < 0.05 after Bonferroni correction using χ2 test between genotypes within conditions (asterisks in (A) and (F)) or between conditions in rme-2(-) animals (hash in (F)) are indicated.

Extent of silencing in progeny by short or mixed dsRNA injected into parental circulation varies between tissues but has similar nuclear Argonaute requirements.

(A-C) GTBP-1::GFP fluorescence from the ubiquitously expressed gene gtbp-1::gfp in the F1 progeny of uninjected P0 animals (no injection) or of P0 animals injected into the body cavity with synthetic 50-bp gfp-dsRNA (short dsRNA) or gfp-dsRNA transcribed from a ∼730-bp DNA template (mixed dsRNA) was analyzed. The expression of gtbp-1::gfp is dimmer in P0 animals (imaged as adults) than in F1 animals (imaged as L4s) because of developmental variation in the expression – therefore comparisons are only appropriate during the same generation and not across generations. (A) Schematic illustrating injection site and scoring scheme. For the soma, a region between the pharynx and anterior gonad arm within a circle (blue, data in (C)) or along a ventral to dorsal (V-D) axis (black, data in (B)) was quantified. For the germline, a gonadal region that excluded the intestine (purple, data in (C)) was quantified. (B) Quantification of F1 progeny after injection of two different concentrations of short dsRNA (1X, 350 ng/μl, left; ∼14X, 4977 ng/μl, right) into the body cavity of P0 animals. (top) The relative mean intensity profile of fluorescence along the V-D axis for progeny of uninjected animals (black), animals injected with short dsRNA (red), or animals injected with mixed dsRNA (blue). Shaded bands indicate 95% CI. (bottom) Ratios of mean intensities within interior points (hashes in top) to those of the exterior points (asterisks in top) are depicted for each imaged animal. (C) Quantification of P0 (black) and F1 (grey) wild-type, nrde-3(tm1116) or hrde-1(tm1200) animals. Regions within the soma and the germline were quantified as indicated in (A). The numbers of P0 and F1 animals quantified (P0; F1 n) are indicated. For each genotype, F1 progeny after no injection, short dsRNA injection, or mixed dsRNA injection into P0 animals showed significantly different fluorescence values from each other (P < 0.05 after Bonferroni correction using Mann-Whitney U test for two-sided comparisons). Similarly significant differences between treatments across genotypes are indicated (asterisks).

In contrast, when synthesized gfp-dsRNA of a defined length (50 bp) with a fluorescent label was injected into circulation in adult animals, no entry into the germline was observed in the absence of RME-2 (Marré et al., 2016). We found that silencing of unc-22 in progeny by similarly synthesized but unlabeled 50-bp unc-22-dsRNA with a 5’ OH delivered into parental circulation also showed a strong requirement for RME-2, unlike mixed dsRNA (Figure 3A).

Further comparison between the two forms of dsRNA revealed that silencing in progeny by 50-bp dsRNA injected into parental circulation was detectably less efficient in somatic cells (Figure 3B and Figure 3–figure supplement 3A and 2B, left), even when ∼14X more 50-bp dsRNA was delivered into parental circulation (Figure 3–figure supplement 3B, right), and was also less efficient in the germline (Figure 3B and Figure 3–figure supplement 3A and 2C). Efficient silencing in response to added dsRNA requires nuclear Argonaute proteins: NRDE-3 in somatic cells (Guang et al., 2008) and HRDE-1 in the germline (Buckley et al., 2012). Both 50-bp dsRNA and mixed dsRNA relied on HRDE-1 for silencing within the germline (Figure 3–figure supplement 3A and 2C) and could silence independent of NRDE-3 in somatic cells (Figure 3– figure supplement 2A and 2C). In addition to the diversity of RNA lengths observed in mixed dsRNA, another known feature that could distinguish dsRNA transcribed in bacteria or in vitro from synthesized 50-bp dsRNA is the presence of 5’ triphosphates on the transcribed dsRNA species instead of the 5’ OH present in synthesized 50-bp dsRNA. In support of the impact of 5’ phosphates on transport and/or silencing, addition of 5’ monophosphates to synthesized 50-bp dsRNA injected into parental circulation reduced the dependence on RME-2 for silencing in progeny (Figure 3–figure supplement 3E and 1F), potentially by enhancing the ability of synthesized dsRNA to be imported by other dsRNA importers (e.g. SID-1) in the absence of RME-2. Thus, the requirements for dsRNA entry into the germline and subsequent silencing vary for different lengths and/or chemical forms of dsRNA (see summary in Table 1).

Fluorescently labeled 50-bp dsRNA delivered into parental circulation localized within intestinal cells in progeny (Figure 3C, top left), as has been observed for vitellogenin proteins (Sharrock, 1983) and fluorescent dyes (Bossinger and Schierenberg, 1996). Accumulation of fluorescently labeled dsRNA was also detected at the apical membrane of the intestine, which could reflect exocytosis of dsRNA into the lumen of developing intestinal cells. However, separation of the fluorescent label from dsRNA catalyzed by cellular enzymes cannot be excluded. Therefore, to dissect differences, if any, between the transport of unlabeled short dsRNA (synthesized 50-bp with 5’OH) and mixed dsRNA (mixture transcribed in vitro using ∼1 kb DNA template) we injected unc-22-dsRNA into animals with mutations in genes that play roles in the import of dsRNA. We found that maternal SID-1 was required for silencing by short dsRNA in progeny (Figure 3C, bottom, left bars), suggesting that the SID-1-dependent entry of short dsRNA into the cytosol likely occurs in the injected parent or during early development in progeny. Uptake of dsRNA from the intestinal lumen requires SID-2, a transmembrane protein located in the apical membranes of intestinal cells (Winston et al., 2007; McEwan et al., 2012). We found that SID-2 was not required for most silencing in progeny by short or mixed dsRNA injected into parental circulation (Figure 3C, top right and bottom). Exit of dsRNA from intracellular vesicles requires SID-5, a transmembrane protein located in endolysosomal membranes (Hinas et al., 2012). Silencing in wild-type animals was comparable to silencing in sid-5(-) animals (Figure 3C, top right). However, when animals that lacked SID-1 were injected, SID-5 was required in progeny for silencing by mixed dsRNA from parental circulation (Figure 3C, bottom, right bars; as also reported in Wang and Hunter, 2017). Since dsRNA is expected to be present in vesicles upon entry through RME-2 in the absence of SID-1 (Marré et al., 2016; Wang and Hunter, 2017), this observation suggests that SID-5 is required for the release of mixed dsRNA from inherited vesicles in progeny.

In summary, injected extracellular dsRNA can enter the germline in parents and be transmitted to progeny through two routes with different substrate selectivity. One route is preferentially used by short dsRNA and relies on RME-2-mediated endocytosis of dsRNA into oocytes, where early exit from vesicles is required for silencing in progeny as evidenced by the need for maternal SID-1 (Figure 3D, blue). The other route appears to exclude short dsRNA but allows mixed dsRNA entry into the cytosol in the parental germline through SID-1 and exit from inherited vesicles in progeny through a process that requires both zygotic SID-1 and SID-5 (Figure 3D, grey; Wang and Hunter, 2017).

Expression of SID-1 is consistent with a role in the intergenerational transport of extracellular dsRNA

All routes of dsRNA transport deduced using the experimental addition of dsRNA ultimately require SID-1 for entry into the cytosol. The proposed model (Figure 3D) for dsRNA transport into the germline and to progeny suggests that the expression pattern of SID-1 is likely to include the germline. We used Cas9-mediated genome editing to insert a piRNA-resistant mCherry sequence (Zhang et al., 2018; Devanapally et al., 2021) into the sid-1 coding sequence (Figure 4A) to observe the endogenous expression pattern of SID-1::mCherry. This fusion protein was detectably functional in contrast to other attempts at tagging SID-1 (see Technical comments in Materials and methods and Figure 4–figure supplement 4). Fluorescence from SID-1::mCherry was visible in the proximal and distal regions of the adult germline (Figure 4B).

The expression pattern of SID-1 varies during development.

(A) Schematic depicting insertion of mCherry sequence that lacks piRNA binding sites (jm195[mCherryΔpi]; Zhang et al., 2018; Devanapally et al., 2021) into the sid-1 gene using Cas9-mediated genome editing. (B and C) Representative images showing fluorescence from SID-1::mCherry (black) in (B) the adult gonad arm, (C, left) developing embryos, (C, middle) L1-stage animals, or (C, right) L4-stage animals with sid-1(jam195[mCherryΔpi]) compared to autofluorescence in wild-type animals of the same stages. Numbers (n) of each stage imaged are indicated (100% of animals exhibited the depicted expression patterns). For animals imaged in (B), the distal germline was obstructed by the intestine in 1/10 sid-1(jam195[mCherryΔpi]) and 5/9 wild-type animals. (D) Airyscan image of an L1-staged animal assembled by stitching four different Z-stacks after depth-coding and taking maximum projections, illustrating the expression of SID-1::mCherry throughout the animal. Scale bar for adult gonad arms in (B) and embryos in (C), 20 μm; scale bar for larvae in (C), 50 μm and in (D), 10 μm. Also see Technical comments on “Making a sid-1 translational reporter” in Materials and methods.

Unsuccessful attempts to functionally tag SID-1 and to identify SID-1-dependent genes.

(A) Schematic illustrating the tagging of sid-1 (box, exon; line, intron) at the 3’ end to generate fusion proteins with fluorophores (GFP, DsRed, or wrmScarlet) tagged at the C-terminus. (B) Images showing subcellularly localized fluorescence (black) from SID-1::DsRed (top) and SID-1::GFP (bottom) within muscle cells when expressed from multicopy transgenes. Scale bar = 10 µm and insets show brightfield images. (C) Structure of SID-1 predicted by AlphaFold shaded based on pLDDT scores (blue/cyan, high; yellow/orange, low). Red arrow indicates the C-terminus. (D) Principal component analysis of RNA-seq experiment comparing transcriptomes from wild-type, sid-1(qt9[nonsense]), sid-1(tm2700[deletion]), and sid-1(tm2700[deletion]); tmIs1005[sid-1(+)] animals. (E) List of sid-1-dependent genes identified by comparing polyA+ RNA from sid-1(qt9[nonsense]) animals with wild-type animals (left) and by comparing sid-1(tm2700[deletion]) animals with sid-1(tm2700[deletion]); tmIs1005[sid-1(+)] animals (right).

Tetracycline-induced functional rescue of sid-1 expression is evident in somatic tissues but not within the germline. (A) Schematic illustrating a cell expressing sid-1 transcript with a tetracycline aptazyme (Wurmthaler et al., 2019) inserted into the sid-1 3’UTR (left) in the presence (bottom right) or absence (top right) of tetracycline. Tetracycline stabilizes sid-1 transcripts by inhibiting ribozyme-based cleavage in the 3’UTR and thereby allows for the expression of SID-1 protein and dsRNA import. (B) Fraction of wild-type or sid-1(jam112[tet]) animals silenced after ingestion of bli-1-dsRNA (left) or expression of neuronal unc-22-dsRNA (right) in the presence of water (grey bars) or 10 μM tetracycline (green bars). Numbers of animals scored for silencing (n) are depicted. (C) The extent of gfp silencing in gtbp-1::gfp; sid-1(jam112[tet]) day 3 adult animals after ingestion of gfp-dsRNA in the presence of water or 10 μM tetracycline. A schematic illustrating the experimental design (top left), representative images of animals from each condition with numbers of animals imaged (n) and percentages of animals represented (bottom left), and quantification of representative germline (top right) and somatic (bottom right) GTBP-1::GFP intensity (a.u.) are depicted. Mean germline GFP intensity was measured in representative regions of the posterior germline and somatic GFP intensity was measured along a dorsal to ventral axis in the tail region (shaded region represents 95% CI) to avoid increased autofluorescence in the intestines of animals exposed to tetracycline. Scale bars, 100 μm. (D) Representative images of gtbp-1::gfp; sid-1[jam112[tet]) F1 day 1 adult animals after P0 and F1 ingestion of gfp-dsRNA until day 1 of F1 adulthood in the presence of different concentrations of tetracycline (10 μM, 20 μM, 50 μM). Higher concentrations of tetracycline did not enhance silencing in gtbp-1::gfp; sid-1(jam112[tet]) animals. Scale bars, 100 μm. (E) The extent of gfp silencing in cross progeny of gtbp-1::gfp; sid-1(jam112[tet]) hermaphrodites injected with water or 10 μM tetracycline and sid-1(jam112[tet]); Ex[rgef-1p::gfp-dsRNA] males in the presence of water or 10 μM tetracycline. A schematic illustrating the experimental design including injection of gtbp-1::gfp; sid-1(jam112[tet]) hermaphrodites with water or 10 μM tetracycline (top left), representative images of animals with the Ex[rgef-1p::gfp-dsRNA] array from each condition with numbers of animals imaged (n) and percentages of animals represented (bottom left), and quantification of representative germline (top right) and somatic (bottom right) GFP intensity (a.u.) as in (C) are depicted. Scale bars, 100 μm. (F) Total brood of wild-type or sid-1(jam112[tet]) animals after culturing on OP50 E. coli or pos-1-dsRNA bacteria in the presence of water or 10 μM tetracycline. Silencing by pos-1-dsRNA typically results in inviable embryos (wild type, bottom), but culturing sid-1(jam112[tet]) with 10 μM tetracycline and pos-1-dsRNA only resulted in a minor decrease in brood size (sid-1(jam112[tet]), bottom). This decrease was not observed when sid-1(jam112[tet]) animals were cultured on 10 μM tetracycline plates in the absence of pos-1-dsRNA (top, brood of 1 animal; bottom, brood of 3 animals). (G) Representative fluorescence images of GTBP-1::GFP (black) in the heads, distal germlines, proximal germlines, and tails of gtbp-1::gfp animals with a tetracycline-aptazyme sequence inserted into the gtbp-1::gfp 3’UTR (gtbp-1(jam210[tet])) after culturing with water or 10 μM tetracycline. The numbers of animals imaged (n) and the percentages of animals with the represented expression patterns are depicted. An increase in GTBP-1::GFP intensity was observed in the soma and germline, but increased fluorescence in the intestine cannot be distinguished from increased autofluorescence caused by culturing with 10 μM tetracycline. Scale bars, 50 μm.

Expression also progressively increased during development with tissue-specific enrichment in the developing embryo (Figure 4C, left), becoming ubiquitous in hatched L1 larvae (Figure 4C, middle, and 4D). SID-1::mCherry was not easily detectable in the germline during later larval development (Figure 4C, middle and right). In combination with the expression of RME-2 observed in the proximal germline (Figure 1–figure supplement 1D; Grant and Hirsh, 1999), this expression pattern of SID-1 is consistent with the entry of most dsRNA from circulation of adult animals into the germline, followed by activity of transport mechanisms in developing embryos and early larvae that inherit parental dsRNA.

To determine if acute, induced expression rather than developmental expression of SID-1 can be sufficient for the import of dsRNA into the germline, we engineered the endogenous sid-1 gene to transcribe a fusion transcript with an aptamer-regulated ribozyme (Figure 4–figure supplement 4A, left) that cleaves itself when not bound to tetracycline (Figure 4–figure supplement 4A, right) (based on Wurmthaler et al., 2019). Exposing these animals to tetracycline enabled silencing by dsRNA in somatic tissues (hypodermis: Figure 4–figure supplement 4B, left; body-wall muscles: Figure 4–figure supplement 4B, right), indicative of stabilization of sid-1 mRNA, production of SID-1 protein, and subsequent dsRNA import in somatic cells.

However, such tetracycline-induced silencing was not detectable in the germline (Figure 4– figure supplement 2C-1F). Yet, similar tagging of the ubiquitously expressed gene gtbp-1::gfp resulted in detectable rescue of expression within the germline by tetracycline (Figure 4–figure supplement 4G). A possible explanation for the poor rescue of SID-1 activity within the germline is that post-transcriptional mechanisms targeting sid-1 mRNA in the germline but not in somatic cells interfere with tetracycline-dependent stabilization of the sid-1 transcript (e.g., piRNA-based regulation of sid-1 mRNA (Ouyang et al., 2019; Dodson and Kennedy, 2019), or that acute stabilization of the sid-1 transcript does not override developmental regulation of SID-1 translation.

Additional attempts to tag the SID-1 protein guided by structure and to modulate sid-1 transcripts guided by post-transcriptional regulatory interactions could improve control of dsRNA transport between cells. Nevertheless, the developmentally regulated expression observed for both SID-1 and RME-2 in the germline is consistent with intergenerational or transgenerational effects of endogenous dsRNA from parental circulation after development of the adult germline.

Temporary loss of sid-1 causes a transgenerational increase in the levels of mRNA from two germline genes

To understand how the dsRNA importer SID-1 might be used in endogenous gene regulation across generations, we searched for sid-1-dependent changes in gene expression that could be heritable (Figure 5, Figure 2–figure supplement 2 and Figure 5–figure supplement 5). To control for genetic background (see Technical comments in Materials and methods), we used Cas9-mediated genome editing to delete the entire sid-1 coding sequence or introduce a nonsense mutation in cohorts of the same wild-type animals. By comparing polyA+ RNA from this wild type with that of the newly generated sid-1(jam113[deletion]) (Figure 5A and 5B and Figure 5– figure supplement 1A) or sid-1(jam80[nonsense]) (Figure 5A-5C) animals, we found that 26 genes were significantly (q < 0.05) misregulated in sid-1(jam113[deletion]) (Figure 5–figure supplement 5B) and 6 in sid-1(jam80[nonsense]) (Figure 5D, top), both including sid-1 (a list of significantly altered genes is in Table 2). The most upregulated gene in sid-1(jam113[deletion]), F14F9.5 (Figure 5–figure supplement 5B), which is located immediately 3’ to sid-1 in the genome, was only misregulated in the deletion mutant and not in the nonsense mutant (Figure 5–figure supplement 5C, left). Both mutants, however, were equally defective for silencing by ingested dsRNA (Figure 5B). This observation suggests that while both mutations result in loss of SID-1 protein, the deletion of sid-1 also changes local regulatory sequences (potentially explaining upregulation of the neighboring gene F14F9.5) and eliminates sid-1 mRNA, which could participate in RNA-based regulatory interactions within the germline (Ouyang et al., 2019; Dodson and Kennedy, 2019). Nevertheless, we could detect two genes that were upregulated in both sid-1(jam113[deletion]) and sid-1(jam80[nonsense]) animals (red in Figure 5D, top, and Figure 5–figure supplement 5B): the identical loci W09B7.2/F07B7.2 (Figure 5–figure supplement 5C, middle), and Y102A5C.36 (Figure 5–figure supplement 5C, right). Intriguingly, another gene cls-3 also changed in both mutants (Table 2) but in different directions (∼3.4-fold decrease in the sid-1(jam80[nonsense]) mutant but a ∼5.8-fold increase in the sid-1(jam113[deletion]) mutant), suggesting that the direction of change in expression can vary.

Ancestral loss of SID-1 causes transgenerational changes in the mRNA levels of two germline genes that are subject to RNA regulation.

(A) Schematic of modifications at the sid-1 gene generated using Cas9-mediated genome editing. Deletion of the entire coding sequence (jam113[deletion]), a nonsense mutation (jam80[nonsense]), and its reversion to wild-type sequence (jam86[revertant]) are depicted. (B) Fractions of animals with the indicated genotypes that show silencing in response to unc-22-dsRNA (grey) or bli-1-dsRNA (black). Numbers of animals scored (n), significant differences using two-tailed test with Wilson’s estimates for single proportions (asterisks, P < 0.05 with Bonferroni correction) and 95% CI (error bars) are indicated. (C) Principal components explaining the variance between wild type (black), sid-1(jam80[nonsense]) (red), and sid-1(jam86[revertant]) (grey) polyA+ RNA samples. Almost all of the variance between samples is explained by PC 1. (D) Volcano plots of changes in the abundance of polyA+ RNA in sid-1(jam80[nonsense]) (top) and sid-1(jam86[revertant]) (bottom) animals compared with wild-type animals (black, q < 0.05; red, both q < 0.05 and change in the same direction in sid-1(jam80[nonsense]) and sid-1(jam113[deletion]); see Figure 5–figure supplement 5). While sid-1 transcript levels in sid-1(jam86[revertant]) are comparable to that in wild type (grey), sdg-1 (W09B7.2/F07B7.2) and sdg-2 (Y102A5C.36) transcript levels remain elevated in sid-1(jam86[revertant]) (red). (E) Levels of spliced sid-1 (top), sdg-1 (middle) and sdg-2 (bottom) transcripts measured using RT-qPCR. The median of three technical replicates is plotted for each of three biological replicates (bar indicates median) assayed before and after 1 year of passaging animals (year 1, dark grey; year 2, light grey). Asterisks indicate P < 0.05 with Bonferroni correction using two-tailed Student’s t-test. (F) Heatmap showing changes in the levels of transcripts (total RNA or mRNA) or antisense small RNAs (22G RNA) from sid-1, sdg-1, sdg-2, and tbb-2 (abundant germline transcript for comparison). Fold changes (expressed as LogFC, indicating log2 for (m)RNA, log10 for piRNA binding, and log10 for 22G RNA) were deduced by integrating reports (study) of 21 experiments that identify subsets of genes as being subject to RNA-mediated regulation within the germline (# genes). These prior studies include comparisons of RNA or 22G RNA from wild-type animals with that from mutant animals (e.g., mut-16(-) 22G RNA), biochemical detection of piRNA binding to transcripts (piRNA-bound mRNA), and biochemical detection of 22G RNA binding to an Argonaute (HRDE-1-bound 22G RNA). ‘NS’ indicates cases where changes, if any, were not significant based on the criteria used in the study. A conservative value of 2-fold is assigned to all genes reported as changing >2-fold in Ni et al., 2016.

Selective disruption of sid-1 followed by restoration to wild type reveals sid-1-dependent transcripts expressed in the germline that show heritable change.

(A) Principal components explaining the variance between wild type (black) and sid-1(jam113[deletion]) (red) animals. (B) Volcano plots of changes in the abundance of polyA+ RNA in sid-1(jam113[deletion]) animals compared with wild-type animals (black, q < 0.05; red, q < 0.05 and with change in the same direction in sid-1(jam80[nonsense]); see Figure 5D, top). (C) Read coverage in biological triplicate (black, blue and purple) at sid-1 and F14F9.5 (left), W09B7.2/F07B7.2 (sdg-1) (represented by F07B7.2 locus, middle) and Y102A5C.36 (sdg-2) (right) of polyA+ RNA in wild-type and sid-1(jam113[deletion]) animals (top), and in wild-type, sid-1(jam80[nonsense]), and sid-1(jam86[revertant]) animals (bottom) normalized to total mapped reads per sample. Deletion of sid-1 coding sequence caused accumulation of transcripts from F14F9.5 (blue), requiring point mutation (jam80[nonsense]) for selective disruption of sid-1 (see Figure 5). (D) Volcano plots of changes in the abundance of RNA in wild-type gonads vs. whole animals (left), mut-16(-) vs. wild-type animals (middle), and prg-1(-) vs. wild-type animals (right) using data from Reed et al., 2020. sdg-1, sdg-2 and sid-1 transcripts are highlighted (red). (E) Levels of spliced sdg-1 and sdg-2 transcripts in animals of the indicated genotypes measured using RT-qPCR. The median (line) of three technical replicates is plotted for each of three biological replicates. P > 0.05 with Bonferroni correction using two-tailed Student’s t-test for wild type to mutant comparisons. Levels of sid-1 transcripts were not detectable in sid-1(jam113[deletion]) animals due to absence of sid-1 coding sequence (data not shown).

Conservatively, we began by analyzing only the two genes with mRNA levels that changed in the same direction in both sid-1 mutants. Both W09B7.2/F07B7.2 and Y102A5C.36 have been reported (Reed et al., 2020) to be expressed within the germline (Figure 5–figure supplement 5D, left) and regulated by endogenous small RNAs (Figure 5–figure supplement 5D, middle and right). Spliced mRNA levels measured at a later generation using RT-qPCR demonstrated that both transcripts were upregulated in sid-1(jam80[nonsense]) animals compared to wild-type animals as expected (Figure 5E), but no upregulation was detectable in sid-1(jam113[deletion]) animals (Figure 5–figure supplement 5E). This difference between the two sid-1 mutants could reflect increased variation in expression (as was observed for cls-3 using RNAseq) or could reflect complex effects caused by deletion of sid-1 DNA (e.g., F14F9.5 overexpression, loss of sid-1 mRNA, etc.) that could be independent of SID-1 protein function.

To determine if changes in W09B7.2/F07B7.2 and Y102A5C.36 expression were heritable, we reverted the sid-1 nonsense mutation to wild-type sequence using Cas9-mediated genome editing. This immediately restored silencing by ingested dsRNA (Figure 5B) with concomitant recovery of sid-1 mRNA to wild-type levels (Figure 5E, top). In contrast, changes in both W09B7.2/F07B7.2 and Y102A5C.36 expression persisted (Figure 5D, bottom) even after one year of passaging the reverted animals (sid-1(jam86[revertant])) (i.e., after >100 generations, Figure 5E, middle and bottom). Thus, the sid-1-dependent accumulation of mRNA from these two germline genes persisted for many generations, likely through mechanisms that maintain heritable epigenetic changes. We hereafter refer to these sid-1-dependent genes (sdg) that show heritable epigenetic changes in response to temporary loss of SID-1 as sdg-1 (W09B7.2/F07B7.2) and sdg-2 (Y102A5C.36).

The sid-1-dependent gene sdg-1 is affected by many factors that regulate RNA silencing in the germline

To determine if expression of sdg-1 and sdg-2 is regulated by other proteins that play a role in RNA silencing within the germline, we examined 21 published datasets (Buckley et al., 2012; Dodson and Kennedy, 2019; Welker et al., 2007; Spike et al., 2008; Wan et al., 2021; Suen et al., 2020; Wahba et al., 2021; Shen et al., 2018; Goh et al., 2014; Lee et al., 2012; Batista et al., 2008; Kim et al., 2021; Ni et al., 2016) that reported changes that depend on such proteins.

For each dataset, we determined if the lists of genes reported as showing significant changes in mutants compared to the respective wild types included sdg-1 and/or sdg-2. This analysis revealed that changes in mRNA and/or antisense small RNAs of sdg-1 were detected in 20 of the 21 datasets while changes in sdg-2 were observed in 9 of 21 (Figure 5F). When detected, changes in sdg-2 were in the same direction as changes in sdg-1, suggestive of similar regulation of both genes.

RNAs transcribed in the germline can be recognized as they exit the nuclear pores by piRNAs bound to the Argonaute PRG-1, which recruits them for regulation by antisense small RNAs called 22G RNA made by proteins organized within perinuclear germ granules (reviewed in Sundby et al., 2021). Interaction with piRNAs was detected for RNA from sid-1, sdg-1, and sdg-2, and the control gene tbb-2 using crosslinking, ligation, and sequencing of hybrids (Shen et al., 2018; Figure 5F), consistent with their germline expression. Depletion of downstream 22G RNAs targeting both sid-1 and sdg-1 was detectable upon loss of the germ granule component MUT-16 (Suen et al., 2020; Figure 5F). Both genes were among the top 500 genes targeted by 22G RNAs bound by the secondary Argonaute HRDE-1/WAGO-9 (Buckley et al., 2012; Figure 5F), suggesting similar downregulation of both genes using 22G RNAs. Furthermore, multiple datasets support downregulation of sdg-1 within the germline by HRDE-1/WAGO-9-bound 22G RNAs in the absence of PRG-1. One, loss of HRDE-1/WAGO-9 increased sdg-1 RNA in whole animals (Ni et al., 2016; Figure 5F) and in dissected gonads (Kim et al., 2021; Figure 5F). Two, loss of PRG-1 decreased sdg-1 RNA (Figure 5F) and increased 22G RNAs that are antisense to sdg-1 (Figure 5F) in dissected gonads (Wahba et al., 2021). Three, although animals that lack PRG-1 become progressively sterile, the increase in sdg-1 22G RNA persisted in near-sterile animals (Figure 5F, near-sterile in Wahba et al., 2021), and this increase was eliminated upon additional loss of HRDE-1/WAGO-9 (Figure 5F, near-sterile in Wahba et al., 2021).

As expected for sid-1-dependent downregulation of sdg-1, multiple datasets support an inverse relationship between the two genes. In animals lacking PRG-1, sid-1 RNA levels increased and sid-1 22G RNAs decreased (Lee et al., 2012; Figure 5F), but both sdg-1 RNA and sdg-2 RNA levels decreased along with an increase in 22G RNAs (Suen et al., 2020; Wahba et al., 2021; Shen et al., 2018; Goh et al., 2014; Lee et al., 2012; Batista et al., 2008; Figure 5F). This inverse relationship between sid-1 and sdg-1 RNA regulation is also observed when many components of germ granules are mutated as indicated by changes in 22G RNA upon loss of the embryonic P granule components MEG-3/-4 (Dodson and Kennedy, 2019; Figure 5F), the PRG-1 interactor DEPS-1 (Suen et al., 2020; Figure 5F), or the Z granule component ZSP-1 (Wan et al., 2021; Figure 5F; also known as PID-2 (Placentino et al., 2021)).

In addition to the above studies, pioneering studies that used microarrays identified sdg-1 as upregulated in animals lacking the germ granule component DEPS-1 (Spike et al., 2008; Figure 5F) and in animals lacking the dsRNA-binding protein RDE-4 (Figure 5F; second-most upregulated in Welker et al., 2007), which recruits dsRNA imported through SID-1 and other intracellular dsRNA for processing and eventual gene silencing. Animals that lack RDE-4 show a ∼47.5-fold increase in sdg-1 RNA (Welker et al., 2007). A reduction in RDE-4 activity could also contribute to the ∼11.6-fold increase in sdg-1 RNA seen in deps-1(-) animals because these animals also show a ∼3.2-fold decrease in rde-4 RNA (one of 13 downregulated genes; Spike et al., 2008). These observations support the idea that appropriate regulation of sdg-1 RNA requires both piRNA-mediated processes that act via germ granule components such as DEPS-1 and dsRNA-mediated processes that use SID-1 and RDE-4.

In summary, the levels of sdg-1 RNA are detectably regulated by the dsRNA-selective importer SID-1, the dsRNA-binding protein RDE-4, and the piRNA-binding Argonaute PRG-1. Presence of dsRNA-mediated regulation or loss of piRNA-mediated regulation enhances MUT-16-dependent production of secondary small RNAs that bind the secondary Argonaute HRDE-1/WAGO-9. Consistent with downregulation of these sid-1-dependent transcripts by SID-1, disruption of many components of germ granules results in opposite effects on these transcripts and sid-1 RNA. Intriguingly, a search of protein interaction studies revealed that the SDG-1 protein is among the interactors of two germ granule components: PID-2 by immunoprecipitation (Placentino et al., 2021; also known as ZSP-1 (Wan et al., 2021)) and DEPS-1 by proximity labeling (Price et al., 2021). This interaction raises the possibility that the sdg-1 gene is not only regulated by germ granule components, but also that its protein product associates with one or more components of germ granules.

Regulation of sdg-1 RNA is susceptible to epigenetic changes that last for many generations

Given these observations on the regulation of sdg-1 and the potential for the SDG-1 protein to be a regulator of RNA silencing, we focus here on the analysis of this gene. The sdg-1 gene is located within a retrotransposon (Figure 6–figure supplement 6A) that is within a duplicated ∼40 kb region and has two recognizable paralogs (Figure 6–figure supplement 6B).

The sdg-1 gene is prone to stochastic changes in gene expression that can become heritable.

(A) Representative images showing fluorescence of SDG-1::mCherry (black) in a wild-type background. While most animals showed symmetric expression in the germline (left), animals with >2-fold difference in fluorescence between both gonad arms (bright anterior, middle and bright posterior, right) were also observed. Punctate fluorescence in the intestine likely represents autofluorescence. Scale bar, 50 μm. (B) Quantification of SDG-1::mCherry fluorescence intensity (arbitrary units, a.u.) in adult gonad arms (anterior arm, dark grey; posterior arm, light grey) of sdg-1(jam137[mCherryΔpi]) animals starting in one generation (x) and continuing in successive generations as indicated. Numbers of gonad arms quantified (n) is indicated. Expression in one generation was not significantly different when compared to that in the previous tested generation using Mann-Whitney U test for two-sided comparisons and Bonferroni correction. (C) Lineages and estimated relative sdg-1 expression 10 generations after mating wild-type (open circle) males with sdg-1::mCherryΔpi (filled circle) hermaphrodites and vice versa, and isolating sdg-1(+) or sdg-1::mCherry animals from F1 heterozygotes (half-filled circle). Expression of sdg-1 in the F10 generation was measured by RT-qPCR of sdg-1 mRNA purified from pooled wild-type animals of mixed stages or by quantification of SDG-1::mCherry fluorescence in gonad arms of adult sdg-1::mCherryΔpi animals. Relative levels of sdg-1 mRNA and SDG-1::mCherry fluorescence intensity were converted to units of estimated relative sdg-1 expression (see Materials and methods) for comparison. See Figure 6–figure supplement 6A for raw data. (D-F) Fluorescence intensity measurements (quantified as in (B)) in adult animals with sdg-1::mCherryΔpi (+) and additionally with mutations in genes introduced through genetic crosses (in regulators of dsRNA import rme-2, sid-2 or sid-5, or in regulators of RNA silencing mut-16 or eri-1) or through genome editing (in regulators of dsRNA import sid-1 or sid-3, or in regulators of RNA silencing rde-1 or deps-1). Asterisks indicate P < 0.05 with Bonferroni correction using Mann-Whitney U test for two-sided comparisons between animals with sdg-1::mCherryΔpi (+) and animals with additional mutations. Nonsense mutations (nonsense) or deletions (deletion) introduced through genetic crosses (isolate numbers #1, #2, etc. in (D)) or genome editing (different alleles in (E) and (F)) and numbers of gonad arms (n) quantified for each isolate are indicated. Mutations in genes required for dsRNA import or subsequent silencing resulted in fewer animals with asymmetric fluorescence between gonad arms (a combined proportion of 21/197 for sid-1, sid-3, rde-1 and deps-1 mutants versus 22/84 for wild type, P = 0.0009 using two-tailed test with Wilson’s estimates for single proportions). Animals with at least one gonad arm brighter than the dimmest wild-type gonad arm in (A) and with asymmetric gonad arms were found in different genotypes (anterior bright: sid-1(-) – 5/122, sid-3(-) – 1/29, rde-1(-) – 2/22, deps-1(-) – 4/24, and posterior bright: sid-1(-) – 6/122, rde-1(-) – 2/22, deps-1(-) – 1/24). (G) Fluorescence intensity measurements as in (B) of animals with sdg-1::mCherryΔpi that show loss of fluorescence when a nonsense mutation is introduced in sid-1 using genome editing ∼30 generations (gen.) later remain changed despite reversion of sid-1 nonsense mutation to wild-type sequence after ∼20 additional generations. Subsequent mutation of deps-1 after another ∼110 generations restored SDG-1::mCherry fluorescence to wild-type levels. Also see Figure 6–figure supplements 1 and 2.

The sid-1-dependent gene sdg-1 is expressed from two identical loci (W09B7.2/F07B7.2) and loss of its expression in sid-1(nonsense) animals fails to recover in sid-1(revertant) animals.

(A) Schematic adapted from UCSC Genome Browser depicting W09B7.2/F07B7.2 (red) located within a repeated ∼40-kb locus on chromosome V (8813207-8854700 depicted; duplicate locus at 8855302-8896495) that includes many histone genes (dark blue; duplicate genes also depicted). W09B7.2/F07B7.2 are located within full-length Cer9 retrotransposons with repeated regions in grey (darker color indicates fewer repeat element-associated mismatches/insertions/deletions). Loci encoding gag and pol elements (PR: protease, RT: reverse transcriptase, RH: RNaseH, IN: integrase) within Cer9 are depicted. (B) Alignment of the SDG-1 protein sequence encoded by W09B7.2/F07B7.2 to the paralogs ZK262.8 and C03A7.2 with conserved residues between two (grey) or three (black) proteins highlighted. (C) Schematic depicting insertion of mCherry sequence that lacks piRNA binding sites (Zhang et al., 2018; Devanapally et al., 2021) at the 3’ end of sdg-1(jam137[mCherryΔpi]), as well as deletion of the sdg-1 coding sequence (jam232, jam241, jam242, jam244, jam245, and jam246). (D) Genotyping gels showing insertion of mCherryΔpi sequences (1095 bp) (left) or deletion of sdg-1 coding sequences (425 bp) (right) at both loci of sdg-1. Absence of wild-type bands in either case confirm genome editing of both copies. (E) Levels of spliced sid-1 (left) and sdg-1 (right) transcripts in wild-type animals and sdg-1(jam137[mCherryΔpi]) animals with a wild-type (+), sid-1(jam150[nonsense]) or sid-1(jam169[revertant]) background measured using RT-qPCR. The median of three technical replicates is plotted for each of three biological replicates (bar indicates median). Asterisks indicate P < 0.05 with Bonferroni correction using two-tailed Student’s t-test.

Mating but not genome editing can initiate distinct heritable changes in sdg-1 expression.

(A) (P0 to F10, top) Quantification of SDG-1::mCherry fluorescence intensity (a.u.) in adult gonad arms (anterior arm, dark grey; posterior arm, light grey) across generations after mating hermaphrodite and male sdg-1(jam137[mCherryΔpi]) animals with male and hermaphrodite wild-type animals, respectively. The generations assayed and numbers of gonad arms quantified (n) are indicated. In F1 and F2, fluorescence intensity values of animals with lineages that were not propagated to F10 but were heterozygous or homozygous sdg-1(jam137[mCherryΔpi]), respectively, were pooled with values of animals with lineages that were propagated to F10. In F3 to F10, top, animals from four different F1 lineages were scored. Fluorescence intensity of animals descending from the self-progeny of P0 sdg-1(jam137[mCherryΔpi]) animals was measured in each generation and is depicted, with the same data plotted for each mating direction for comparison. Asterisk indicates P < 0.05 with Bonferroni correction using Mann-Whitney U test for two-sided comparisons. (F10, bottom) Levels of spliced sdg-1 mRNA transcripts in wild-type animals, sdg-1(jam137[mCherryΔpi]) animals and two lineages of wild-type F10 progeny from each cross direction, measured using RT-qPCR. The median of three technical replicates is plotted for each of three biological replicates (bar indicates median). Asterisks indicate P < 0.05 with Bonferroni correction using two-tailed Student’s t-test. (B) (P0 and F1) Schematic illustrating mutation of dpy-10 in three P0 lineages of sdg-1(jam137[mCherryΔpi] animals and subsequent segregation of the dpy-10 mutation. (F2 and F3) Both dpy-10(-) and dpy-10(+) F2 and F3 animals from each of the three P0 lineages were imaged and SDG-1::mCherry intensity was quantified (a.u.) in adult gonad arms (anterior arm, dark grey; posterior arm, light grey). Minor differences in SDG-1::mCherry expression were observed between mutants and nonmutants in some cases, as well as between lineages. The numbers of gonad arms quantified (n) are depicted. Asterisks indicates P < 0.05 with Bonferroni correction using Mann-Whitney U test for two-sided comparisons.

However, there are no detectable SDG-1 homologs in species other than C. elegans, suggesting that it is a recently evolved gene. Potentially because of this lack of conservation, AlphaFold is unable to predict a confident structure for SDG-1 (CELE_F07B7.2 on the Database (Varadi et al., 2022)). To facilitate analysis of SDG-1 expression, we tagged both loci that encode SDG-1 with mCherry coding sequences lacking piRNA-binding sites (mCherryΔpi; Zhang et al., 2018; Devanapally et al., 2021) (Figure 6–figure supplement 6C and 1D). This tagging resulted in the expression of sdg-1::mCherryΔpi mRNA being ∼16-fold higher than sdg-1 mRNA (Figure 6– figure supplement 1E), potentially because of the reduction in the overall density of piRNA-binding sites per transcript, the additional introns included in mCherryΔpi (based on Okkema et al., 1993 and Crane et al., 2019), and/or other unknown factors. Fluorescence from SDG-1::mCherry was observed in the germline of adult animals (Figure 6A). However, animals showed variation in SDG-1::mCherry expression between their two gonad arms (Figure 6A, middle shows bright anterior (20% of animals) and right shows bright posterior (6% of animals)). A contributing feature for the observed stochasticity could be the location of sdg-1 within a duplicated region (Figure 6–figure supplement 6A), as suggested by similar stochastic RNA silencing of multi-copy transgenes but not single-copy transgenes (Le et al., 2016). Despite this variation, unbiased passaging of self-progeny for more than 18 generations continuously preserved SDG-1::mCherry expression in an otherwise wild-type background (Figure 6B). In contrast, mating, which can perturb RNA regulation within the germline in cross progeny (Devanapally et al., 2021), caused dramatic changes in sdg-1 expression that persisted in descendants (Figure 6C). Mating animals that express SDG-1::mCherry with wild-type animals resulted in heritable changes along lineages that express sdg-1::mCherryΔpi mRNA or that express sdg-1 mRNA (Figure 6C and Figure 6–figure supplement 6A). This discovery of mating-induced perturbation in gene expression raises caution in interpreting past studies (summarized in Figure 5F) where changes in multiple independent isolates were not examined. Nevertheless, when we used genetic crosses to determine the impact of mutations on sdg-1 expression, we observed reduced SDG-1::mCherry fluorescence in mutants predicted to have reduced levels of intracellular dsRNA (Figure 6D, sid-2(-), sid-5(-)) or reduced processing of intracellular dsRNA (Figure 6D, eri-1(-)). In contrast, we observed an increase in SDG-1::mCherry fluorescence in animals lacking MUT-16 (Figure 6D). Finally, animals lacking RME-2, which lack the ability to import many maternal factors (e.g., lipids, proteins, RNAs, etc.), also showed an increase in SDG-1::mCherry fluorescence (Figure 6D).

To avoid mating-induced perturbations of RNA regulation within the germline, we used Cas9-mediated genome editing to introduce mutations into animals that express SDG-1::mCherry in an otherwise wild-type background. Use of this approach to mutate a control gene with no known roles in RNA regulation within the germline resulted in similar levels of SDG-1::mCherry fluorescence in multiple isolates of animals with and without the mutation (Figure 6– figure supplement 2B). In contrast, mutating sid-1 using Cas9-mediated genome editing caused a range of expression levels in different isolates when compared with sid-1(+) animals (Figure 6E - 6 isolates lower, 2 isolates comparable, and 1 isolate higher), which differs from the increase in sdg-1 mRNA observed upon SID-1 loss in the single isolate examined earlier (Figure 5). Mutating sid-3 also lowered the levels of SDG-1::mCherry in one isolate, but caused no detectable change in another (Figure 6F). While both isolates with loss of RDE-1 showed lower levels of SDG-1::mCherry, both isolates with loss of the germ granule component DEPS-1 showed higher levels of SDG-1::mCherry (Figure 6F). These experiments also reveal that in most isolates of animals expected to have reduced levels of intracellular dsRNA (sid-3(-), sid-1(-)) or reduced processing of intracellular dsRNA (rde-1(-)), SDG-1::mCherry fluorescence is reduced. In contrast, isolates lacking DEPS-1 showed increased expression of SDG-1::mCherry.

Collectively, the observations on SDG-1 expression using mutants suggest that the uptake and processing of intracellular dsRNA (influenced by SID-1, SID-2, SID-3, SID-5, RDE-1, and ERI-1) and the function of germ granules (influenced by MUT-16 and DEPS-1) are both necessary to maintain intermediate levels of SDG-1 expression across generations. Once the levels of SDG-1::mCherry were reduced upon loss of SID-1, downregulation persisted across generations even after restoration of wild-type SID-1 (Figure 6G), just as the upregulation of untagged sdg-1 mRNA also persisted (Figure 5). Despite >100 generations of such persistent silencing, the expression of SDG-1::mCherry could be restored by mutating deps-1 (Figure 6G), implicating small RNA-based regulation and germ granules in the maintenance of the new epigenetic states established upon loss of SID-1.

SID-1-dependent genes, including SDG-1, could alter RNA-mediated regulation in the germline

Since SDG-1 interacts with regulators of RNA silencing (ZSP-1/PID-2 (Placentino et al., 2021) and DEPS-1 (Price et al., 2021)), loss of SID-1 could both block the entry of extracellular dsRNA into the cytosol and change intracellular RNA regulation through SID-1-dependent genes such as SDG-1. In support of this possibility, intracellular delivery of dsRNA through injection into the syncytial germline of sid-1(-) animals showed a weak defect in silencing (Figure 2D in Marré et al., 2016 and Figure 1 in Wang and Hunter, 2017). To examine if changes in the levels of SDG-1 alone could account for such a defect in silencing by intracellular dsRNA in the germline, we exposed independently-generated sdg-1 deletion animals and animals that overexpress sdg-1 (i.e., with sdg-1::mCherryΔpi) to dsRNA matching the germline gene pos-1 for ∼16 hours. This short exposure to pos-1-dsRNA caused intermediate levels of silencing in wild-type animals and comparable intermediate silencing in sdg-1 mutant and sdg-1 overexpressing animals (Figure 7A), suggesting that changes in sdg-1 levels alone are not sufficient to cause a defect in silencing by dsRNA that is detectable using this assay.

SID-1 modifies RNA regulation within the germline, potentially through sdg-1 and other sid-1-dependent genes.

(A) (left) Schematic of assay for sensitive detection of pos-1 silencing by ingested dsRNA. (right) Numbers of developed progeny (> 3rd larval stage) laid by wild-type animals, animals with a deletion (Δ) in sdg-1 (jam232, jam241, jam242) or animals with overexpression (sdg-1::mCherryΔpi) of sdg-1 exposed to pos-1 dsRNA (red) or control dsRNA (black) for 16 hours are plotted. Asterisks indicate P < 0.05 using Mann-Whitney U test for two-sided comparisons with Bonferroni correction. (B) Cross progeny males that inherited the mex-5p::mCherry::h2b::gfp::h2b transgene (T; Devanapally et al., 2021; also used in Figure 2) from maternal (left) or paternal (right) parents, both of wild-type, sid-1(-), or sdg-1(-) background, were scored for expression of mCherry and GFP (bright, dim, off). Wild-type data for top set (n = 77 and n = 33) are replotted from Devanapally et al., 2021 for comparison. Dashed line separates independent experiments. Asterisk indicates P < 0.05 with Bonferroni correction using χ2 test; n.s. indicates not significant. (C) Representative AiryScan images of the distal germline (left; scale bar, 10 µm) or single germline nuclei (right; scale bar, 2 µm) showing SDG-1::mCherry alone (top) or with GFP::ZNFX-1 (bottom, merge and single channel images). The number of animals imaged (n) and the percentage that show enrichment of SDG-1::mCherry in perinuclear foci are indicated. Sites of SDG-1::mCherry enrichment coincide with GFP::ZNFX-1 localization. Boxes in left mark the nuclei shown in right. (D) Representative images showing entry of SDG-1::mCherry into the nucleus in −1 oocytes (left) and upon pronuclear fusion in early embryos during the time course indicated (right). Numbers of germlines and embryos imaged are indicated. Scale bars, 20 µm. Also see Videos 1-4. (E) Representative image of the hermaphrodite germline in animals with a translational (left) or transcriptional (right) reporter of sdg-1. Scale bars, 20 µm. Apparent extracellular punctae of SDG-1::mCherry and mCherry surrounding the proximal germline requires further study, but could be non-specific because similar localization is observed in animals with other promoters driving mCherry expression, but not GFP expression, in the germline (data not shown). The numbers of animals with the depicted fluorescence pattern are indicated. (F and G) Response of the transcriptional sdg-1 reporter (sdg-1p::mCherryΔpi[sdg-1(Δ)]::sdg-1 3’ UTR) to the addition of unc-22-dsRNA (F) or loss of rde-4 (G). Quantification and asterisk are as in Figure 6. (H) Models for dsRNA import into the germline (top) and subsequent RNA-mediated regulation of sdg-1 (bottom). See text for details.

Alternatively, the previously described defect in silencing by intracellular dsRNA in sid-1(-) animals (Marré et al., 2016; Wang and Hunter, 2017) could be through the promotion of competing piRNA-mediated gene regulation in the absence of SID-1. This notion that dsRNA-mediated and piRNA-mediated gene regulation compete in the germline is supported by a study that demonstrated that loss of PRG-1 enhances heritable RNA silencing by dsRNA (Shukla et al., 2021). To test if loss of SID-1 or a sid-1-dependent gene enhances piRNA-mediated silencing, we used an experimental system which initiates piRNA-mediated silencing of the two-gene operon described in Figure 2 through mating, independent of externally provided dsRNA (Devanapally et al., 2021). We found that sid-1(-) animals exhibited enhanced mating-induced silencing (Figure 7B, top right: ∼50% off in sid-1(+) vs 100% off in sid-1(qt9[nonsense])), while animals lacking sdg-1 showed a small reduction in mating-induced silencing that was not statistically significant (Figure 7B, bottom right, ∼40% off in sdg-1(+) vs ∼30% off in sdg-1(jam232[deletion])). Taken together, these results support the model that loss of SID-1 weakly inhibits silencing by intracellular dsRNA (Marré et al., 2016; Wang and Hunter, 2017) but enhances silencing by piRNAs within the germline.

RNA regulation within the germline using piRNAs relies on phase-separated granules (reviewed in Dodson and Kennedy, 2020). To determine if the previous identification of SDG-1 as a putative interactor of the Z-granule component PID-2 (Placentino et al., 2021)/ZSP-1 (Wan et al., 2021) and potentially of the P-granule-adjacent protein DEPS-1 (Price et al., 2021) could be seen as colocalization with Z-granules in vivo, we examined the localization of SDG-1::mCherry within the cytoplasm at high resolution using AiryScan imaging (Huff et al., 2017). SDG-1::mCherry was enriched in perinuclear foci in many animals (Figure 7C, top; 7 of 9 animals) and these sites of enrichment overlapped with the Z-granule marker GFP::ZNFX-1 (Ishidate et al., 2018; Wan et al., 2018; Figure 7C, bottom; 100% colocalized in 10 of 12 animals with enrichment). However, given the resolution of our imaging, we cannot distinguish between a specific interaction with one of the many Z granule components (Zhao et al., 2024) and an interaction with components of the many other nearby perinuclear granules that continue to be reported (Huang et al., 2024). Therefore, we conservatively conclude that SDG-1::mCherry colocalizes with perinuclear germ granules. Time-course imaging revealed re-localization of SDG-1::mCherry into the nucleus from the cytoplasm in the −1 oocyte, which will be the first to be fertilized (Figure 7D, left and Video 1), along with subsequent exclusion from the nucleus before the maternal and paternal pronuclei meet (Figure 7D, right and Video 2). Such dynamic entry into the nucleus followed by exclusion from the nucleus also occurred during early cell divisions in the developing embryo (Videos 3 and 4). The timing of nuclear entry of SDG-1::mCherry coincides with the nuclear envelope breakdown events that occur during fertilization and early development (Huelgas-Morales and Greenstein, 2018). The sdg-1 coding sequence was required for regulated nuclear entry because deletion of the sdg-1 open reading frame in sdg-1::mCherryΔpi animals resulted in mCherry expression throughout the germline in both the cytoplasm and nuclei (Figure 7E). Nuclear localization of the SDG-1 protein in the −1 oocyte is like that of the essential Argonaute CSR-1b (Charlesworth et al., 2021), thought to play a role in protecting transcripts from silencing. Exposure to ingested dsRNA did not alter the expression of sdg-1p::mCherryΔpi[sdg-1(Δ)]::sdg-1 3’ UTR (Figure 7F) but loss of rde-4 perturbed expression such that one isolate showed loss of expression while the other showed enhanced expression (Figure 7G). These results suggest that while the sdg-1 open reading frame is required for exclusion from germline nuclei, it is not required for the response of the sdg-1 gene to changes in intracellular dsRNA. Together, these observations on the levels and localizations of SDG-1 raise the possibility that this protein is actively regulated by extracellular and intracellular dsRNA with a role in heritable RNA silencing, potentially with perinuclear germ granules, and additional cell-cycle coupled roles during early development, potentially through nucleocytoplasmic shuttling.

Discussion

Our analyses suggest a model for the impact of dsRNA from parental circulation on descendants (Figure 7H). Extracellular dsRNA can accumulate in parental circulation through regulated secretion from neurons (e.g., oxidative damage promotes accumulation) and potentially other tissues. Uptake into the germline is RME-2-dependent for some forms of dsRNA (blue in Figure 7H), but RME-2-independent for other forms of dsRNA (grey in Figure 7H). While all forms of dsRNA require SID-1 for the entry of dsRNA into the cytosol in progeny, RME-2-dependent dsRNA also require SID-5. Such dsRNA from parental circulation along with other intracellular dsRNA are processed with the help of the dsRNA-binding protein RDE-4 in progeny (Figure 7H, bottom). This processing of dsRNA regulates the SID-1-dependent gene sdg-1 by reducing variation between animals in the levels of sdg-1 mRNA, although the dsRNA sequences likely need not match sdg-1 because no such dsRNA was reported in the published dsRNAome (Whipple et al., 2015) and because this regulation can occur in the absence of the sdg-1 open reading frame. The sdg-1 gene is located within a retrotransposon that is also targeted by heritable RNA silencing, but the SDG-1 protein is enriched near perinuclear germ granules (observed using the Z granule component ZNFX-1), which are required for heritable RNA silencing. Consistent with the SDG-1 protein promoting heritable RNA silencing, the sdg-1::mCherryΔpi gene is highly susceptible to mating-induced silencing, potentially owing to the ∼16-fold higher levels of SDG-1::mCherry compared with that of SDG-1 in wild-type animals. In agreement with this proposal, SID-1 limits heritable RNA silencing because loss of SID-1 enhances mating-induced silencing (Figure 7B, top). Since the sdg-1 gene is located within a retrotransposon that is targeted by heritable RNA silencing, this mechanism for regulating the regulators of heritable RNA silencing such as SDG-1 (and potentially other SDGs) reveals a strategy for tuning an autoregulatory loop for heritable RNA silencing by using competing dsRNA processing. Intriguingly, SDG-1 becomes enriched within nuclei upon nuclear envelope breakdown during fertilization and during early cell divisions in embryos (Figure 7H, bottom right) with active exclusion from nuclei after each reformation of the nuclear envelope, suggestive of additional roles for this retrotransposon-encoded protein.

Import of extracellular dsRNA

The temporal and/or spatial preferences observed for silencing by both ingested dsRNA (Figure 1) and neuronal dsRNA (Figure 2) could be because of unknown characteristics of the exported neuronal dsRNA or ingested dsRNA expressed from bacteria (e.g., modifications, lengths, structures, etc.) that influence import or subsequent silencing. This hypothesis is supported by the different genetic requirements reported for silencing by neuronal gfp-dsRNA compared to other sources of gfp-dsRNA (Ravikumar et al., 2019). Alternatively, these preferences could reflect universal constraints for silencing using any extracellular dsRNA (e.g. expression patterns of factors that promote the import or processing of dsRNA). Furthermore, for both neuronal and ingested dsRNA it is unknown if additional extracellular molecules impact the uptake of dsRNA into cells. Regardless of the presence of such unknown molecules that could be released from neurons to a different degree upon oxidative damage, the gene regulatory effects of the released neuronal dsRNA provide convenient readouts that can be analyzed to understand neuronal damage and its consequences in animals. Pathologies of the central nervous system in humans, including cancer, stroke, multiple sclerosis, neurodegenerative disease, and brain injury, have been associated with extracellular RNAs detected in circulation (reviewed in Tielking et al., 2019), although their origins and regulatory consequences, if any, remain unknown. While the physiological conditions that promote secretion of dsRNA from C. elegans neurons are not known, the discovery that oxidative damage of neurons can enhance the secretion of dsRNA suggests that disruption of cell structures by oxidative damage (e.g., membrane integrity) or initiation of cellular processes that repair oxidative damage (e.g., through ejection of damaged macromolecules (Melentijevic et al., 2017)) also promote the release of dsRNA.

The trafficking of extracellular dsRNA from parent to progeny has spatial specificity, as evidenced by more silencing within the proximal germline (Figure 2), temporal specificity, as evidenced by the need for dsRNA beyond the fourth larval stage (Marré et al., 2016; Wang and Hunter, 2017; Figure 1), and substrate specificity, as evidenced by the differential requirements for 50-bp dsRNA with 5’ OH versus a mix of longer dsRNAs with 5’ triphosphates (Figure 3). The observed difference in the extent of silencing by these dsRNAs delivered into parental circulation cannot be attributed to differential engagement of the nuclear Argonautes HRDE-1 and NRDE-3 in progeny (Figure 3–figure supplement 3), but rather could be the result of differences in the ability of each type of dsRNA to bind to upstream factors in the RNA interference pathway (e.g. RDE-4), differences in the stability of each type of dsRNA, and/or differences in the intergenerational transport of each type of dsRNA. Drivers of temporal specificity during development include changes in the uptake of dsRNA into the intestine, distribution of dsRNA to other tissues from the intestine, import of dsRNA into the germline, and availability of RNA silencing factors within the germline. Proteins mediating dsRNA transport or subsequent silencing also differ in their availability during development and in their affinities for different substrates. For example, SID-1 was not detectable in the developing larval germline but was detected in the adult germline (Figure 4) and has an extracellular domain that binds dsRNA (Li et al., 2015) but could prefer dsRNA molecules with 5’ phosphates and/or long dsRNA. Another consideration for substrate specificity is the composition of ‘mixed dsRNAs’ generated through in vitro transcription versus defined short dsRNAs. In addition to a variety of lengths, mixed dsRNAs could also include a variety of undefined secondary structures. Such structures could recruit or titrate away RNA-binding proteins in addition to the dsRNA structures engaging the canonical RNAi pathway, resulting in mixed mechanisms of silencing. Additionally, although the selectivity uncovered here could apply to all dsRNA delivered into the extracellular space of C. elegans from any source (see Table 1 for constraints on intergenerational dsRNA transport), the chemistry of the delivered dsRNA could be modified by yet unidentified enzymes in vivo to overcome these requirements. Tracking labeled dsRNA with diverse chemistries from parental circulation to progeny could allow correlation of differences observed in progeny silencing to differences in intergenerational trafficking.

Physiological role(s) of SID-1

The germline is a major site of dsRNA import in C. elegans as evidenced by three key observations: the expression of SID-1 in the germline (Figure 4), heritable misregulation of germline genes in sid-1(-) animals (Figures 5 and 6), and accumulation of fluorescently labeled dsRNA from the extracellular space in the germline (Marré et al., 2016; Wang and Hunter, 2017). As a result, sid-1(-) animals could have a defect in the germline that is detectable only under conditions that promote dsRNA transport (e.g., oxidative damage). Multiple physiological defects in the germline and soma of sid-1(-) animals have been reported but have not been widely reproduced, have only been characterized within single generations, or have not been attributed to any specific sid-1-dependent gene(s). These include defects in animals with some misfolded proteins in the endoplasmic reticulum (Long et al., 2014), in animals exiting the dauer stage (Ow et al., 2018; Kadekar and Roy, 2019), in animals exposed to pathogenic P. aeruginosa (Palominos et al., 2017; Moore et al., 2019; Kaletsky et al., 2020), in animals exposed to an odorant (Posner et al., 2019), in intestinal cells that develop in the presence of a multi-copy transgene (Ohno et al., 2022), and in animals that overexpress α-synuclein (Gaeta et al., 2022). RNA-seq experiments in this study and comparisons to those of previous studies suggest that genetic background-dependent changes can obscure genuine sid-1-dependent changes (see Figure 5–figure supplement 5 and Technical comments in Materials and methods), raising caution in the interpretation of putative sid-1-dependent defects. Comparing multiple sid-1 mutants generated using genome editing with animals in which the mutated sequence has been reverted to wild-type sequence in the same genetic background could provide a firmer basis for the identification of additional sid-1-dependent processes while avoiding the potential for mating-induced changes in gene regulation (Devanapally et al., 2021) that can result when genetic crosses are used instead of genome editing.

Retrotransposon-encoded protein regulators

Genes expressed within the germline are likely regulated by positive feedback loops that are required to continually produce factors for maintaining germline immortality and for preserving form and function across generations (Jose, 2020; Chey and Jose, 2021). Thus, germline genes could be particularly vulnerable to heritable epigenetic changes, where deviations in the expression levels of a gene that is regulated by or is part of such feedback loops has the potential to become permanent in descendants. Prior data on sdg-1 mRNA (Figure 5F) and the SDG-1 protein (Placentino et al., 2021; Price et al., 2021) suggested the hypothesis that reduction of sdg-1 RNA via SID-1 alters the amount of SDG-1 protein, which could interact with components of germ granules to mediate RNA regulation within the germline of wild-type animals. Our analysis of sdg-1 expression suggests that it is part of a regulatory architecture that is susceptible to heritable epigenetic changes through the perturbation of RNA regulation (Figures 5-7). Such architectures within the germline could be exploited by ‘selfish’ genetic elements such as retrotransposons to persist across evolution if one of these elements also include genes encoding a regulator. In support of a wider use of such a strategy, a paralog of SDG-1, ZK262.8 (Figure 6–figure supplement 6B), is also encoded by a gene located within a retrotransposon and its loss along with that of the miRNA-associated Argonaute ALG-2 was reported to be synthetic lethal (Tops et al., 2006).

Buffering of heritable epigenetic change by extracellular dsRNA

Loss of SID-1 weakly inhibits silencing by intracellular dsRNA (Marré et al., 2016; Wang and Hunter, 2017) but enhances the initiation of mating-induced silencing within the germline (Figure 7B). However, loss of SDG-1 alone does not account for these effects (Figure 7A and 7B), suggesting that additional SID-1-dependent genes could be involved. Furthermore, the ultimate consequence of SID-1 loss could be different in different isolates as exemplified by the differential changes in cls-3 in sid-1(jam80[nonsense]) and sid-1(jam113[deletion]) animals (Table 2). Indeed, the levels of SDG-1 also vary in different isolates of sid-1(-) animals (Figure 6), suggesting that in the absence of SID-1 the sdg-1 gene becomes prone to heritable epigenetic changes, resulting in either high or low expression states being stabilized in different isolates.

Together these observations suggest that SID-1-dependent import of extracellular dsRNA could play a role in buffering against heritable epigenetic changes in the germline. Such buffering could be necessary to maintain levels of gene expression within a particular range for each regulatory context. Given the enrichment of SDG-1 protein with perinuclear germ granules (Figure 7C) and the maintenance of heritable changes in sdg-1::mCherryΔpi expression by a DEPS-1-dependent process (Figure 6G), buffering against changes in gene expression could involve both RNA- and protein-based regulation that tunes the function of perinuclear germ granules. We therefore speculate that one role for extracellular RNAs that enter germ cells in other systems (e.g., tRNA fragments in mammals (Sharma et al., 2016; Chen et al., 2016; Sharma et al., 2018)) could be to similarly buffer against heritable changes in gene expression.